---
title: "YvO Proteomics Pipeline — Phase 3: Differential Expression"
author: "DTL"
date: "2026-02-25"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# 1. Overview

This phase performs differential expression analysis on the normalized,
**non-imputed** proteomics matrix using `limma` with repeated-measures
modeling via `duplicateCorrelation()`.

**Design:** 2 x 2 factorial (Age x Time) with repeated measures on subject.

**Key decision:** The primary analysis uses non-imputed data because `limma`
handles missing values natively (per-protein `na.rm`), avoiding imputation
artifacts (Karpievitch et al. 2019, BMC Bioinform 20:391). An imputation
sensitivity analysis is run separately to quantify how imputation affects
results.

**Input:** `01_normalization/c_data/02_normalized.csv` (2,124 proteins x 62
samples, 11.67% missing)

**Output:** Per-contrast results tables, combined results, diagnostic plots,
and robustness analyses.

**Script:** `03_DEP/a_script/YvO_DEP_analysis.R`

---

# 2. Statistical Model

## 2.1 Design Matrix

The model uses a cell-means (no-intercept) parameterization with four group
levels:

```
~ 0 + group + (1 | subject)
```

where `group` has levels: `Young_Pre`, `Young_Post`, `Old_Pre`, `Old_Post`
and `subject` is the repeated-measures blocking factor (31 unique subjects).

**Critical implementation note:** The blocking factor uses
`sub("_(Pre|Post)$", "", Col_ID)` rather than `Subject_ID`, because
`Subject_ID` is **not globally unique** — 8 IDs are shared between
Young and Old cohorts (different individuals), and 2 IDs are shared
within the Old group between supplement arms. See the 00_input inventory
for details.

## 2.2 Within-Subject Correlation

`limma::duplicateCorrelation()` estimates the intra-block (within-subject)
correlation, which quantifies how much two measurements from the same person
resemble each other relative to between-person variation.

**Result: r = 0.282** — moderate positive correlation, meaning ~28% of the
variance in protein abundance is attributable to consistent individual
differences. This is typical for proteomics studies with paired pre/post
designs (Smyth 2005, Stat Appl Genet Mol Biol).

**Effect on power:** The within-subject correlation reduces the effective
standard deviation for paired contrasts (Training_Young, Training_Old) by a
factor of $\sqrt{2(1-r)}$, improving power to detect training effects.

**Reference:** limma User's Guide Section 9.7; Smyth 2005.

## 2.3 Contrasts

```{r}
#| label: contrast-matrix
#| eval: false
contrasts_vector <- c(
  "Training_Young = Young_Post - Young_Pre",
  "Training_Old = Old_Post - Old_Pre",
  "Aging = Old_Pre - Young_Pre",
  "Interaction = (Old_Post - Old_Pre) - (Young_Post - Young_Pre)",
  "Reversal = (Old_Post - Old_Pre) - (Old_Pre - Young_Pre)"
)
```

| Contrast | Interpretation | Type |
|----------|---------------|------|
| Training_Young | Proteins changed by training in young adults | Paired (within-subject) |
| Training_Old | Proteins changed by training in older adults | Paired (within-subject) |
| Aging | Baseline age differences (old vs young at pre) | Between-subject |
| Interaction | Age x Time interaction (does training differ by age?) | Between-subject |
| Reversal | Does old post-training move toward young pre-training? | Composite |

## 2.4 Significance Criteria

Three complementary criteria are applied:

1. **FDR < 0.10** (BH-adjusted): Standard multiple-testing correction. The
   relaxed threshold (vs 0.05) is justified by small sample size (~15/group)
   and exploratory proteomics context (Tyanova et al. 2016).
2. **P < 0.01** (nominal): Raw p-value threshold for secondary analysis.
3. **Pi-score < 0.05**: Combined significance and effect size metric (Xiao et
   al. 2014, Bioinformatics). Pi = p^|logFC|; values < 0.05 require both
   statistical significance and meaningful fold-change.

---

# 3. Results

## 3.1 DEP Counts

```{r}
#| label: dep-counts
da <- read.csv("../c_data/02_DA_summary.csv")
knitr::kable(da[, c("contrast", "type", "sig.FDR", "sig.Pi", "sig.Nominal.01")],
             caption = "Differentially expressed proteins per contrast")
```

**Key findings:**

- **Training_Young:** 308 FDR-significant proteins (152 up, 156 down) —
  robust training response.
- **Training_Old:** 0 FDR-significant proteins — not a bug; reflects
  attenuated training response with age (Robinson et al. 2022; Bechshoft
  et al. 2024). The Pi-score identifies 17 proteins with combined
  significance + effect size.
- **Aging:** 405 FDR-significant proteins — largest effect, consistent with
  age being the dominant source of variation.
- **Interaction:** 1 FDR-significant protein — expected given small sample
  size and the subtle nature of age x time interactions.
- **Reversal:** 146 FDR-significant proteins — evidence for partial reversal
  of aging-associated changes after training.

## 3.2 Partial NA Coefficients

11 proteins had partial NA coefficients, meaning they had missing values in
specific group×timepoint combinations that prevented estimation of all
contrasts. These proteins are tested in contrasts where sufficient data exists
and return NA for others.

---

# 4. Robustness Enhancements

## 4.1 Blunting Diagnostics (Hypothesis 1)

**Question:** Is the training response truly attenuated in older adults, or
is the difference between Training_Young and Training_Old just noise?

**Tests applied:**

- **Kolmogorov-Smirnov test** on |logFC| distributions: D = 0.181,
  p = 1.4 x 10^-30. The distributions are significantly different — young
  adults show systematically larger absolute fold-changes.
- **Fligner-Killeen test** for variance homogeneity: chi-sq = 215.1,
  p = 1.0 x 10^-48. Training_Young has significantly greater variance in
  |logFC|, consistent with a broader, more robust training response.

```{r}
#| label: blunting-results
blunt <- read.csv("../c_data/06_blunting_diagnostics.csv")
knitr::kable(blunt, digits = 4, caption = "Blunting diagnostics")
```

**Report:** `00_DEP_overview.pdf` — composite multi-page PDF including
|logFC| density overlay, bootstrap forest plot, and imputation sensitivity.

## 4.2 Bootstrap Effect Size CIs

```{r}
#| label: bootstrap-results
boot <- read.csv("../c_data/07_effect_size_bootstrap.csv")
knitr::kable(boot, digits = 4, caption = "Median |logFC| with 95% BCa CI")
```

**Interpretation:** The 95% bootstrap CIs for Training_Young [0.132, 0.146]
and Training_Old [0.087, 0.098] do not overlap, confirming that the
attenuation is robust and not driven by outliers.

**Report:** See `00_DEP_overview.pdf` (forest plot page).

## 4.3 Power Analysis

```{r}
#| label: power-results
pwr <- read.csv("../c_data/08_power_analysis.csv")
knitr::kable(pwr, digits = 4, caption = "Minimum detectable logFC at 80% power")
```

**Interpretation:**

- **Paired contrasts** (Training_Young, Training_Old): The within-subject
  correlation (r = 0.282) reduces effective sigma from 0.667 to 0.566,
  giving minimum detectable logFC of ~0.37–0.38 at 80% power. This means
  the study can detect ~30% fold-changes for training effects.
- **Unpaired contrasts** (Aging, Interaction): Minimum detectable logFC of
  ~0.62, meaning the study can detect ~54% fold-changes for age differences.
- **Implication for Training_Old:** The zero FDR hits are consistent with
  Training_Old having genuinely smaller effects (median |logFC| = 0.092)
  that fall below the detection threshold (~0.38).

## 4.4 Imputation Sensitivity

```{r}
#| label: sensitivity-results
sens <- read.csv("../c_data/09_imputation_sensitivity.csv")
knitr::kable(sens, digits = 4,
             caption = "Spearman correlation of t-statistics: non-imputed vs imputed")
```

**Interpretation:** Spearman rho ranges from 0.932 to 0.976 across contrasts
— near-perfect rank agreement. This means:

1. The **rank ordering** of proteins by significance is essentially unchanged
   by imputation.
2. Downstream analyses (GSEA, clustering) that depend on protein rankings
   will give concordant results regardless of imputation choice.
3. The primary (non-imputed) analysis is well-supported.

**Report:** See `00_DEP_overview.pdf` (imputation sensitivity page) —
scatter plots of t-statistics with significance-switching proteins highlighted.

---

# 5. Output Files

## 5.1 Data Outputs (`c_data/`)

| File | Description |
|------|-------------|
| `01_limma_DAList.rds` | Saved DAList object for downstream figure scripts |
| `02_DA_summary.csv` | Counts of DEPs per contrast and criterion |
| `03_combined_results.csv` | All proteins with logFC, t, P, adj.P, pi-score per contrast |
| `04_per_contrast_results/*.csv` | Individual contrast result tables |
| `05_results.xlsx` | Excel workbook with one sheet per contrast + summary |
| `06_blunting_diagnostics.csv` | KS + Fligner test results |
| `07_effect_size_bootstrap.csv` | Median |logFC| with 95% BCa CIs |
| `08_power_analysis.csv` | Minimum detectable logFC at 80% power |
| `09_imputation_sensitivity.csv` | Spearman rho (non-imputed vs imputed) |
| `10_DEP_supplementary.xlsx` | Supplementary Excel workbook (methods, per-contrast, robustness) |

## 5.2 Report Outputs (`b_reports/`)

| Report | Contents |
|--------|----------|
| `00_DEP_overview.pdf` | Multi-page composite: DEP counts, blunting density, bootstrap forest, imputation sensitivity |
| `{contrast}/summary.pdf` | Triple volcano (nominal, FDR, Pi-score) + top 25 table |
| `{contrast}/*.pdf` | Standard limma diagnostic plots (MD, volcano, p-hist) |
| `{contrast}/*_DA_report.html` | Interactive HTML report (proteoDA) |

---

# 6. Known Limitations

1. **Training_Old = 0 FDR hits:** This is biologically plausible (attenuated
   response with age) and consistent with the power analysis showing minimum
   detectable logFC exceeds the observed median effect size. Robinson et al.
   2022 and Bechshoft et al. 2024 report similar attenuation.

2. **>20% DA warning:** proteoDA warns that Training_Young, Aging, and
   Reversal contrasts have >20% nominally significant proteins, potentially
   violating the assumption that most proteins are not DA. This is common
   in exercise proteomics where training produces widespread remodeling
   (Hostrup et al. 2022). The empirical Bayes shrinkage in limma is
   somewhat robust to this violation.

3. **fGSEA not cross-corrected:** Pathway enrichment (run in figure scripts)
   is independent per contrast without cross-contrast correction. This is
   standard but should be acknowledged in the manuscript methods.

4. **Interaction power:** Only 1 FDR-significant interaction protein reflects
   the inherent power limitation for detecting interaction effects with
   ~15 subjects per group.

---

# 7. Downstream Dependencies

| Consumer | Input | What It Uses |
|----------|-------|-------------|
| F2 (Volcanos, RRHO, fGSEA) | `01_limma_DAList.rds`, `03_combined_results.csv` | Training_Young, Training_Old ranked lists |
| F3 (Reversal) | `01_limma_DAList.rds`, `03_combined_results.csv` | Aging + Training_Old for reversal tests |
| F4 (Clustering) | `04_per_contrast_results/` | Identifies DEPs to subset for clustering |
| F5 (WGCNA) | `03_combined_results.csv` | Module-trait correlations use DEP statistics |
| F6 (Classification) | `03_combined_results.csv` | Feature selection from DEP signatures |

---

# 8. Session Information

```{r}
#| label: session-info
sessionInfo()
```
