---
title: "YvO Proteomics Pipeline — Phase 1: Normalization"
author: "DTL"
date: "2026-02-26"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# 1. Overview

This phase transforms raw DIA-MS protein intensities into a normalized,
quality-controlled expression matrix suitable for downstream differential
expression analysis. The pipeline performs:

1. **Tissue relevance filtering** — Removal of proteins not annotated in
   skeletal muscle by the Human Protein Atlas
2. **Missingness-based filtering** — Retention of proteins detected in ≥66%
   of samples within at least one experimental group
3. **Outlier detection** — Consensus approach using three independent methods
4. **Normalization** — Cyclic loess (cycloess) via the proteoDA package

**Input:** `00_input/YvO_raw.xlsx` (2,837 proteins × 64 samples) +
`00_input/YvO_meta.xlsx` (sample metadata)

**Output:** `01_normalization/c_data/02_normalized.csv` (2,124 proteins × 62
samples) + 4 diagnostic PDF reports

**Script:** `01_normalization/a_script/YvO_normalization_run.R`

---

# 2. Literature & Methods Justification

## 2.1 Normalization Method: Cyclic Loess

**Cyclic loess** (cycloess) iteratively applies loess regression between all
pairs of samples to remove systematic intensity-dependent biases. It was chosen
based on:

- **Bolstad et al. 2003** (Bioinformatics 19:185–193): Foundational comparison
  of normalization methods for high-density arrays, demonstrating that
  loess-based approaches outperform global methods for intensity-dependent
  biases. The cyclic loess implementation in `limma` builds on this work.
- **Välikangas et al. 2018** (Brief Bioinform 19:1–11): Systematic evaluation
  of normalization methods for label-free proteomics. Cycloess consistently
  ranked among top performers across multiple evaluation criteria.
- **Arend et al. 2025** (Brief Bioinform 26:bbaf201, PRONE package):
  Large-scale normalization evaluation showing that the optimal method is
  dataset-specific — supporting the use of `proteoDA::write_norm_report()` to
  compare methods on this specific dataset.

**Why not other methods?**

| Method | Reason for rejection |
|--------|---------------------|
| Quantile normalization | Too aggressive for heterogeneous samples; can distort biological variation between groups |
| Median centering | Only corrects global shifts, not intensity-dependent biases |
| VSN | Performs well but assumes homoscedastic variance, which may not hold for extreme intensities |
| LOESS (non-cyclic) | Only pairwise; cyclic version ensures transitivity |

The `proteoDA::write_norm_report()` function generates a multi-method
comparison report (saved as `01_norm_comparison.pdf`) so the choice can be
empirically validated on this specific dataset.

## 2.2 Tissue Relevance Filtering (HPA)

The **Human Protein Atlas** (Uhlén et al. 2015, Science 347:1260419) provides
tissue-specific expression and subcellular location annotations. Filtering to
HPA-annotated skeletal muscle proteins removes likely contaminants (plasma
proteins, proteins from non-muscle tissues in the biopsy).

**Rationale:** DIA-MS is comprehensive and will detect proteins present even at
low abundance. Some of these may be contaminants from blood, connective tissue,
or sample handling. Restricting to HPA-annotated proteins improves
signal-to-noise for muscle-biology questions.

**Sensitivity safeguard:** The script includes an "HPA sensitivity audit" that
checks whether any aging-relevant proteins (collagens, myosin heavy chains,
HLA, IGF, MSTN, PAX7, FOXO, atrogenes, mitochondrial, antioxidant, cytokines)
were inadvertently removed.

## 2.3 Missingness Filtering (66% Group Rule)

A protein is retained if it is detected in ≥66% of samples within **at least
one** experimental group (Young_Pre, Young_Post, Old_Pre, Old_Post).

**Why 66%?**
- Balances coverage vs. information loss
- Group-aware: a protein present in all Young but missing in all Old is still
  biologically informative (potential age-dependent expression)
- Consistent with recommendations from O'Leary et al. 2025

**Why group-aware (not global)?**
- A global threshold (e.g., ≥50% of all samples) would remove proteins
  exclusively expressed in one group, potentially discarding the most
  interesting biology.

## 2.4 Outlier Detection (Consensus Approach)

Three independent methods flag outlier samples:

1. **Paired missingness** — Excessive missing data or large missingness
   difference between Pre/Post pairs (IQR + 1.5×IQR threshold)
2. **PCA Mahalanobis distance** — Samples far from the multivariate centroid in
   PC1–PC3 space (χ² cutoff at p < 0.01 with df = 3)
3. **MAD-based median intensity** — Samples whose median log2 intensity
   deviates >3 MAD from the global median

A sample is flagged only if **all three** methods agree (3/3 consensus). When
one timepoint of a subject is flagged, both Pre and Post are removed to
maintain the paired design.

**Reference:** Huang et al. 2024 (Brief Bioinform 25:bbae129, SEAOP) describes
ensemble outlier approaches for proteomics, supporting multi-method consensus.

## 2.5 Coefficient of Variation on Linear Scale

CV is calculated on **linear-scale intensities** directly from `dal$data`
(which is already on the linear scale, median ~2.1M), not on a log2 scale.

**Reference:** Brenes 2024 (J Proteome Res 23:5274–5278, PMC11629372): The
standard CV formula (SD/mean) must be applied to linear-scale intensities, not
log-transformed values. Applying SD/mean to log2 data artificially compresses
CV by up to 14-fold, producing misleadingly low values.

---

# 3. Data Flow

```
YvO_raw.xlsx (2837 proteins × 64 samples)
    │
    ├── HPA tissue filter → 2672 proteins (-165)
    ├── Deduplication → 2670 proteins (-2)
    ├── 66% group missingness → 2124 proteins (-546)
    │
    ├── Outlier detection (3-method consensus)
    │   └── Y_S05 flagged → remove Pre + Post → 62 samples
    │
    └── Cycloess normalization → 2124 proteins × 62 samples
```

---

# 4. Code Walkthrough

## 4.1 Setup & Package Loading

```{r}
#| label: setup
#| eval: false

pacman::p_load(
  proteoDA,                         # DAList object + normalization
  readxl, readr, dplyr, tidyr, stringr,
  org.Hs.eg.db, GO.db, AnnotationDbi,  # GO:CC annotations
  ggplot2, ggrepel, patchwork, openxlsx
)

set.seed(42)                         # Reproducibility
setwd(rprojroot::find_rstudio_root_file())  # Project root
```

**What this does:** Loads all required packages via `pacman` (which installs
missing packages automatically). Sets a random seed for reproducibility and
changes to the project root directory.

**Key package:** `proteoDA` (Thurman et al. 2023, JOSS 8:5184) provides the
`DAList` object structure, normalization methods, and QC report generation.

## 4.2 Data Loading

```{r}
#| label: data-loading
#| eval: false

raw <- read_excel(file.path(input_dir, "YvO_raw.xlsx"))
annotation <- raw[, c("uniprot_id", "protein", "gene", "description", "n_seq")]
intensity  <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_excel(file.path(input_dir, "YvO_meta.xlsx")))
stopifnot("Sample mismatch" = setequal(colnames(intensity), metadata$Col_ID))
```

**What this does:** Reads the raw intensity Excel file, separates annotation
columns (protein identifiers) from intensity columns (one per sample). Reads
the metadata and verifies that all sample column names match metadata Col_IDs.

**Result:** 2,837 proteins × 64 samples.

## 4.3 HPA Tissue Filter

```{r}
#| label: hpa-filter
#| eval: false

hpa <- read_tsv(file.path(input_dir, "HPA_skeletal_muscle_annotations.tsv"))
keep_mask <- annotation$gene %in% hpa$Gene
intensity  <- intensity[keep_mask, ]
annotation <- annotation[keep_mask, ]
```

**What this does:** Loads the Human Protein Atlas skeletal muscle gene list
(12,893 genes). Keeps only proteins whose gene symbol appears in the HPA list.

**Result:** 2,837 → 2,672 proteins (165 removed, 5.8% loss). The GO:CC donut
chart in the filtering report shows the subcellular compartments of removed
proteins.

## 4.4 Deduplication

```{r}
#| label: dedup
#| eval: false

# If multiple rows share the same UniProt ID, keep the one with highest mean intensity
annotation$row_mean <- rowMeans(data.matrix(intensity), na.rm = TRUE)
keep_idx <- annotation %>%
  group_by(uniprot_id) %>%
  slice_max(row_mean, n = 1, with_ties = FALSE) %>%
  pull(row_idx)
```

**What this does:** In rare cases, the same protein (UniProt ID) may appear
multiple times due to different peptide-to-protein groupings. The highest-
abundance version is retained.

**Result:** 2,672 → 2,670 proteins (2 duplicates removed).

## 4.5 Missingness-Based Filtering

```{r}
#| label: missingness-filter
#| eval: false

group_prop <- dal$metadata %>%
  split(.$Group_Time) %>%
  lapply(function(g) rowMeans(!is.na(dal$data[, g$Col_ID, drop = FALSE])))

# Keep protein if detected in ≥66% of samples in ANY group
keep <- apply(group_prop >= 0.66, 1, any)
dal <- filter_proteins_by_annotation(dal, keep)
```

**What this does:** For each experimental group (Young_Pre, Young_Post,
Old_Pre, Old_Post), calculates the proportion of samples in which each protein
is detected (not NA). A protein passes if its detection rate reaches 66% in at
least one group.

**Why 66%?** This threshold ensures that a protein is present in a clear
majority of samples within the group where it's most consistently detected.
This is more permissive than requiring 66% globally (which would eliminate
group-specific proteins) but more stringent than no filtering (which would
retain proteins detected in only 1–2 samples).

**Result:** 2,670 → 2,124 proteins (546 removed, 25.7% of this step). These
are predominantly low-abundance proteins with sparse detection across all
groups.

## 4.6 CV Variability Assessment

```{r}
#| label: cv-assessment
#| eval: false

# CV on linear-scale intensities (Brenes 2024) — dal$data is already linear
lin_dat <- dal$data

cv_by_group <- dal$metadata %>%
  split(.$Group_Time) %>%
  lapply(function(g) {
    sub <- lin_dat[, g$Col_ID, drop = FALSE]
    apply(sub, 1, function(x) sd(x, na.rm=TRUE) / mean(x, na.rm=TRUE))
  })
```

**What this does:** Calculates the coefficient of variation (CV = SD/mean) for
each protein within each group. CV is computed on linear-scale intensities (not
log2) per Brenes 2024. The data in `dal$data` is already on the linear scale
(median ~2.1M), so no back-transformation is needed.

**What the output shows:** The CV violin plot and scatter plots in the
diagnostics report reveal inter-individual variability. Pre vs Post CV scatter
plots should fall near the 1:1 line (no systematic change in variability over
time). Young vs Old comparisons show whether one age group has higher
variability.

## 4.7 Outlier Detection

```{r}
#| label: outlier-detection
#| eval: false

# Method 1: Paired missingness
pct_missing <- colMeans(is.na(dal$data)) * 100
miss_threshold  <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing)

# Method 2: PCA Mahalanobis distance
mahal_dist <- mahalanobis(pc_scores, colMeans(pc_scores), cov(pc_scores))
pca_flags  <- mahal_dist > qchisq(0.99, df = 3)

# Method 3: MAD-based median intensity
mad_flags <- abs(sample_medians - global_median) > 3 * mad_val

# Consensus: flagged if ALL 3 agree
consensus_outlier <- (miss_flag + pca_flag + mad_flag) >= 3
```

**What this does:** Three independent methods each flag potentially aberrant
samples. A sample is removed only if all three methods agree, which provides
high specificity (low false-positive rate for outlier calling).

**Result:** 1 subject flagged (Y_S05). Both Pre and Post samples removed to
maintain the paired design. 64 → 62 samples.

**What to look for in the diagnostics report:**
- **Panel A** (Paired Missingness): Flagged samples appear in the upper-right
  quadrant (high missingness + high delta missingness)
- **Panel B** (PCA): Flagged samples (red) should be clear spatial outliers
- **Panel C** (MAD): Flagged samples deviate from the horizontal band

## 4.8 Normalization

```{r}
#| label: normalization
#| eval: false

# Generate comparison report (saved as PDF)
write_norm_report(dal, grouping_column = "Group_Time",
                  output_dir = report_dir, filename = "01_norm_comparison.pdf")

# Apply chosen method
dal <- normalize_data(dal, norm_method = "cycloess")
```

**What this does:** First generates a multi-method comparison report
(`01_norm_comparison.pdf`) that evaluates multiple normalization approaches
on this specific dataset. Then applies cyclic loess normalization.

**What the normalization report shows:**
- Box plots of sample distributions before/after each method
- PCA colored by experimental group
- Intra-group CVs and PCV/PMAD/PEV metrics
- The proteoDA report helps confirm that cycloess is appropriate for this
  dataset

**What the QC reports show:**
- `02_qc_pre.pdf`: Sample distributions, PCA, and hierarchical
  clustering **before** normalization
- `03_qc_post.pdf`: Same plots **after** normalization. Samples within
  the same group should cluster more tightly; systematic biases (batch effects,
  run-order effects) should be reduced.

---

# 5. Output Files

## 5.1 Data Outputs (`c_data/`)

```{r}
#| label: output-listing

outputs <- data.frame(
  File = c("01_DAList_prenorm.rds", "02_normalized.csv",
           "03_DAList_normalized.rds", "04_norm_quality_scores.csv",
           "05_normalization_supp.xlsx"),
  Description = c(
    "DAList snapshot before normalization (for comparison/sensitivity check)",
    "Normalized expression matrix (2124 × 66: 4 annot + 62 samples)",
    "Full R DAList object with metadata, annotations, and normalized data",
    "PRONE-style normalization method ranking (consumed by 03_Sensitivity_check)",
    "Supplementary workbook: Pipeline_Summary, Normalization_Ranking, Outlier_Diagnostics, HPA_Audit, Filtered_Proteins"
  )
)
knitr::kable(outputs, caption = "Data output inventory")
```

## 5.2 Report Outputs (`b_reports/`)

| Report | Contents |
|--------|----------|
| `01_norm_comparison.pdf` | Multi-method normalization comparison (box plots, PCA, CV metrics) |
| `02_qc_pre.pdf` | Pre-normalization QC (distributions, PCA, clustering) |
| `03_qc_post.pdf` | Post-normalization QC (same metrics, should show improvement) |
| `04_diagnostics.pdf` | 3 pages: filtering (donuts + bars + missingness), variance (CV + eta-squared + within/between + PCA), outlier diagnostics |

---

# 6. Results & Interpretation

## 6.1 Protein Retention

```{r}
#| label: filter-summary

filter_log <- openxlsx::read.xlsx("../c_data/05_normalization_supp.xlsx",
                                   sheet = "Pipeline_Summary", startRow = 4)
knitr::kable(filter_log, caption = "Protein retention through filtering pipeline",
             digits = 1)
```

**Interpretation:** Starting from 2,837 raw proteins, 74.9% (2,124) are
retained after all filtering steps. The largest loss is from missingness
filtering (546 proteins), which removes proteins that were only sporadically
detected. This retention rate is typical for DIA-MS skeletal muscle proteomics
(cf. Hostrup et al. 2022: 3,168 → ~2,500 after filtering; O'Leary et al.
2025: similar retention rates).

## 6.2 Outlier Detection

```{r}
#| label: outlier-summary

outlier_diag <- openxlsx::read.xlsx("../c_data/05_normalization_supp.xlsx",
                                     sheet = "Outlier_Diagnostics", startRow = 4)

# Summary
n_flagged <- sum(outlier_diag$consensus_outlier, na.rm = TRUE)
cat(sprintf("Samples flagged by consensus: %d\n", n_flagged))
cat(sprintf("Flagged subjects: %s\n",
    paste(unique(outlier_diag$prefix[outlier_diag$consensus_outlier]),
          collapse = ", ")))
```

**Interpretation:** Only 1 subject (Y_S05) was flagged across all 3 methods,
indicating a genuinely aberrant sample rather than a borderline case. Removing
both Pre and Post samples maintains the paired design (important for
`duplicateCorrelation` in the DEP phase).

## 6.3 Final Dataset

The normalized dataset contains **2,124 proteins × 62 samples** with the
following design:

| Group | Pre | Post | Subjects |
|-------|-----|------|----------|
| Young | 16 | 16 | 16 |
| Old | 15 | 15 | 15 |

Missing values remain in the normalized matrix (15,371 cells, ~11.7% of all
values). These will be addressed in Phase 2 (Imputation).

---

# 7. Validation & Sanity Checks

## 7.1 Expected Behavior

- **PCA (post-norm):** Samples should partially separate by Age (PC1 or PC2),
  with Pre and Post overlapping within each age group. Complete separation by
  timepoint would be unexpected (exercise effects are smaller than age effects
  in proteomics).
- **Box plots (post-norm):** Sample-level intensity distributions should be
  well-aligned after cycloess, with no systematic shifts between groups.
- **CV scatter:** Pre vs Post CV should track the 1:1 line. If Post CVs are
  systematically higher, this could indicate treatment-induced heterogeneity.

## 7.2 Limitations

1. **HPA filter may be overly conservative** — Some biologically relevant
   proteins may lack HPA skeletal muscle annotation. The sensitivity audit
   (HPA_Audit sheet in `05_normalization_supp.xlsx`) quantifies this risk.

2. **Cycloess assumes most proteins unchanged** — This is a standard assumption
   for normalization methods. If the training intervention caused widespread
   directional changes (all proteins up or down), cycloess would partially
   "normalize away" this global shift. However, exercise-induced proteome
   changes are typically bidirectional and affect a minority of proteins,
   making this assumption reasonable.

3. **Outlier removal is irreversible** — Y_S05's removal reduces sample size
   by ~3%. The 3/3 consensus threshold is conservative (minimizes false
   positives) but could miss moderate outliers flagged by only 1–2 methods.

---

# 8. Downstream Dependencies

The normalized DAList (`03_DAList_normalized.rds`) is the primary input for:

- **Phase 2 (Imputation):** Uses the normalized matrix + missingness patterns
- **Phase 3 (DEP):** Uses the normalized matrix directly for limma analysis
- **All figure scripts:** Load metadata from this DAList

**Critical note:** The DAList stores `Subject_ID` in its metadata, but
`Subject_ID` is **not globally unique** (see Phase 0 documentation). Downstream
scripts that need subject-level pairing (e.g., `duplicateCorrelation`) must
derive a unique subject key from `Col_ID`:
```r
subject_key <- sub("_(Pre|Post)$", "", Col_ID)
```

---

# 9. Session Information

```{r}
#| label: session-info
sessionInfo()
```
