---
title: "YvO Proteomics Pipeline — Phase 1: Normalization"
author: "DTL"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
bibliography: ../../references.bib
---

# Overview

This tutorial walks through the first stage of the YvO proteomics analysis pipeline: **normalization**. We start with raw protein intensity data from a DIA-MS experiment and produce a clean, normalized expression matrix ready for downstream imputation and differential expression analysis.

Our study uses a **2x2 repeated-measures design**: Young (n = 16) versus Old (n = 15) participants, each measured before (Pre) and after (Post) a resistance training intervention. DIA-NN library-free search initially identified 2,837 proteins across 64 samples. By the end of this tutorial, we will have a quality-controlled matrix of **2,124 proteins x 62 samples**.

The pipeline performs four major steps:

1. **Tissue relevance filtering** — Remove proteins not annotated in skeletal muscle
2. **Missingness-based filtering** — Keep only proteins reliably detected across samples
3. **Outlier detection** — Flag and remove aberrant samples using a multi-method consensus
4. **Normalization** — Apply cyclic loess to correct systematic intensity biases

> **Note:** This document mirrors the logic in `01_run_normalization.R` and `02_norm_reports.R`. All computations, thresholds, and outputs are preserved exactly as they appear in those canonical scripts. We are adding narrative and context, not changing the analysis.

---

## Prerequisites

### Required packages

```{r}
#| label: setup

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(proteoDA)
library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
```

> **Note:** **proteoDA** [@Thurman2023] is an R package purpose-built for quantitative proteomics. It provides the `DAList` data structure (a container that keeps your intensity matrix, sample metadata, and protein annotations together), normalization methods, and automated QC report generation. We chose it because it integrates tightly with limma and simplifies the normalization-comparison workflow.

### Input files

We need three files from the `00_input/` directory:

| File | Description |
|------|-------------|
| `YvO_raw.xlsx` | Raw protein intensities (2,837 proteins x 64 samples) |
| `YvO_meta.xlsx` | Sample metadata (Col_ID, Group, Timepoint, Subject_ID, Group_Time) |
| `HPA_skeletal_muscle_annotations.tsv` | Human Protein Atlas skeletal muscle gene list |

### Directory structure

All file paths are relative to the project root (`A_YvO_2025/`). We use `rprojroot` to find it automatically, so the code works regardless of where you open RStudio.

```{r}
#| label: config

set.seed(42)

cfg <- list(
  # Input files
  raw_file  = "00_input/YvO_raw.xlsx",
  meta_file = "00_input/YvO_meta.xlsx",
  hpa_file  = "00_input/HPA_skeletal_muscle_annotations.tsv",

  # Output directories
  report_dir = "01_normalization/b_reports",
  data_dir   = "01_normalization/c_data",

  # Thresholds
  miss_pct    = 0.66,       # min detection rate in at least one Group_Time
  outlier_k   = 3,          # methods that must agree for consensus
  mahal_p     = 0.01,       # PCA Mahalanobis chi-sq cutoff
  mad_k       = 3,          # MAD multiplier for median intensity
  norm_method = "cycloess"
)

dir.create(cfg$report_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(cfg$data_dir,   recursive = TRUE, showWarnings = FALSE)
```

We also define a helper function for PCA that we will reuse throughout the pipeline. It performs median imputation **only for the PCA computation** — imputed values are never exported.

```{r}
#| label: helper-pca

run_pca <- function(mat, metadata, log_transform = TRUE) {
  # Median-impute for PCA only (imputed values never exported)
  for (j in seq_len(ncol(mat)))
    mat[is.na(mat[, j]), j] <- median(mat[, j], na.rm = TRUE)
  if (log_transform) mat <- log2(mat + 1)
  pca <- prcomp(t(mat), center = TRUE, scale. = TRUE)
  ve  <- round(summary(pca)$importance[2, 1:3] * 100, 1)
  pc  <- as.data.frame(pca$x[, 1:3]) |>
    mutate(Col_ID = rownames(pca$x)) |>
    left_join(metadata, by = "Col_ID")
  list(pca = pca, scores = pc, var_exp = ve)
}
```

---

## Step 1: Load Raw Data

We begin by reading the raw intensity file and splitting it into two parts: **annotation columns** (protein identifiers like UniProt ID, gene symbol, and description) and **intensity columns** (one column per sample, containing raw DIA-MS intensities).

```{r}
#| label: load-data

raw <- read_excel(cfg$raw_file)

annot_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
annotation <- raw[, annot_cols]
intensity  <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_excel(cfg$meta_file))
rownames(metadata) <- metadata$Col_ID

# Verify that sample columns match metadata
stopifnot("Sample mismatch" = setequal(colnames(intensity), metadata$Col_ID))
intensity <- intensity[, metadata$Col_ID]  # ensure consistent column order

n_raw <- nrow(annotation)
filter_log <- tibble(step = "Raw input", n_before = NA_integer_,
                     n_after = n_raw, n_removed = NA_integer_)
cat(sprintf("Raw: %d proteins x %d samples\n", n_raw, ncol(intensity)))
```

We start with `r n_raw` proteins across `r ncol(intensity)` samples. The `stopifnot()` call is a safety check — it halts execution if the sample columns in the intensity matrix don't exactly match the metadata. This prevents subtle downstream errors from misaligned data.

---

## Step 2: HPA Tissue Filter

Not every protein detected by DIA-MS is relevant to skeletal muscle biology. Some are contaminants from blood, connective tissue, or sample handling. We filter to proteins annotated in the **Human Protein Atlas** (HPA) skeletal muscle database [@Uhlen2015].

```{r}
#| label: hpa-filter

hpa <- read_tsv(cfg$hpa_file, show_col_types = FALSE) |>
  select(Gene, Ensembl, Evidence,
         Protein_class    = `Protein class`,
         Subcellular_main = `Subcellular main location`,
         Interactions) |>
  distinct(Gene, .keep_all = TRUE)

n_before <- nrow(annotation)
keep_hpa   <- annotation$gene %in% hpa$Gene
intensity  <- intensity[keep_hpa, ]
annotation <- annotation[keep_hpa, ] |> left_join(hpa, by = c("gene" = "Gene"))
removed_genes <- setdiff(raw$gene, annotation$gene)

filter_log <- bind_rows(filter_log, tibble(
  step = "HPA tissue filter", n_before = n_before,
  n_after = nrow(annotation), n_removed = n_before - nrow(annotation)))
cat(sprintf("HPA: %d -> %d (-%d)\n",
            n_before, nrow(annotation), n_before - nrow(annotation)))
```

The HPA filter removes `r n_before - nrow(annotation)` proteins (`r round((n_before - nrow(annotation)) / n_before * 100, 1)`% of the starting set). These are proteins whose gene symbols do not appear in the HPA skeletal muscle annotation. While DIA-MS is comprehensive enough to detect them, they are likely not central to our muscle-biology questions.

> **Note:** This filter may be slightly conservative — some biologically relevant proteins could lack HPA skeletal muscle annotation. The supplementary workbook (exported later) lists all removed proteins so you can check for false exclusions.

---

## Step 3: Deduplicate by UniProt ID

In rare cases, the same protein (identified by UniProt ID) may appear in multiple rows because of different peptide-to-protein groupings during the DIA-NN search. When this happens, we keep the row with the highest mean intensity across all samples.

```{r}
#| label: dedup

if (any(duplicated(annotation$uniprot_id))) {
  n_before_dup <- nrow(annotation)
  annotation$row_mean <- rowMeans(data.matrix(intensity), na.rm = TRUE)
  keep_idx <- annotation |>
    mutate(row_idx = row_number()) |>
    group_by(uniprot_id) |>
    slice_max(row_mean, n = 1, with_ties = FALSE) |>
    pull(row_idx)
  annotation <- annotation[keep_idx, ]
  intensity  <- intensity[keep_idx, ]
  annotation$row_mean <- NULL
  filter_log <- bind_rows(filter_log, tibble(
    step = "Deduplication", n_before = n_before_dup,
    n_after = nrow(annotation), n_removed = n_before_dup - nrow(annotation)))
  cat(sprintf("Deduplicated: %d proteins\n", nrow(annotation)))
}
```

This step typically removes only 1--2 proteins. It ensures that each UniProt ID appears exactly once in our matrix, which is required for correct downstream joins and statistical modeling.

---

## Step 4: Assemble DAList and Apply Missingness Filter

Now we assemble our data into a **DAList** — proteoDA's core data structure that bundles the intensity matrix, protein annotations, and sample metadata into a single object. We then filter proteins based on their **detection rate** across experimental groups.

### What is a DAList?

A `DAList` is simply a named list with three slots:

- `$data` — the intensity matrix (proteins x samples)
- `$annotation` — a data frame of protein-level information (UniProt ID, gene name, etc.)
- `$metadata` — a data frame of sample-level information (group, timepoint, subject ID, etc.)

All downstream proteoDA functions (normalization, QC reports, limma fitting) expect this structure.

### Missingness filtering

DIA-MS data contains **missing values** — positions where a protein was not detected in a particular sample. A protein with too many missing values provides unreliable information. We apply a **group-aware 66% detection rule**: a protein is retained if it is detected in at least 66% of samples within **at least one** experimental group (Young_Pre, Young_Post, Old_Pre, or Old_Post).

```{r}
#| label: missingness-filter

int_mat <- as.data.frame(data.matrix(intensity))
rownames(int_mat) <- annotation$uniprot_id
annot_df <- as.data.frame(annotation); rownames(annot_df) <- annotation$uniprot_id
meta_df  <- as.data.frame(metadata);   rownames(meta_df)  <- metadata$Col_ID

dal <- DAList(data = int_mat, annotation = annot_df, metadata = meta_df)
dal <- zero_to_missing(dal)  # convert 0 intensities to NA (DIA-NN reports 0 for undetected)

n_before <- nrow(dal$data)
# For each group, compute the detection rate for every protein
group_prop <- dal$metadata |>
  split(dal$metadata$Group_Time) |>
  lapply(function(g) rowMeans(!is.na(dal$data[, g$Col_ID, drop = FALSE]))) |>
  bind_cols()

# Keep if detection rate >= 66% in ANY group
keep <- apply(group_prop >= cfg$miss_pct, 1, any)
dal  <- filter_proteins_by_annotation(dal, keep)

filter_log <- bind_rows(filter_log, tibble(
  step = sprintf("Missingness (>=%d%% in >=1 group)", cfg$miss_pct * 100),
  n_before = n_before, n_after = nrow(dal$data),
  n_removed = n_before - nrow(dal$data)))
filter_log <- filter_log |> mutate(pct_of_raw = round(n_after / n_raw * 100, 1))
cat(sprintf("Missingness: %d -> %d (-%d)\n",
            n_before, nrow(dal$data), n_before - nrow(dal$data)))
```

We also record which proteins were removed and why, for the supplementary workbook:

```{r}
#| label: record-filtered

filtered_proteins <- bind_rows(
  tibble(uniprot_id = raw$uniprot_id, gene = raw$gene,
         description = raw$description) |>
    filter(gene %in% removed_genes) |>
    mutate(removal_step = "HPA tissue filter"),
  annot_df |>
    filter(!uniprot_id %in% rownames(dal$data)) |>
    select(uniprot_id, gene, description) |>
    mutate(removal_step = "Missingness (<66% in all groups)")
) |> distinct(uniprot_id, .keep_all = TRUE)
```

**Why 66% and why group-aware?** A global threshold (e.g., "detected in 50% of all samples") would remove proteins that are expressed exclusively in one age group — precisely the proteins most likely to show age-related differential expression. The group-aware approach preserves these biologically informative proteins. The 66% value ensures a clear majority of samples within the most-populated group detected the protein, providing enough data points for reliable statistical estimation.

After this step, we have `r nrow(dal$data)` proteins remaining — a `r round(nrow(dal$data) / n_raw * 100, 1)`% retention rate from the raw input.

---

## Step 5: Outlier Detection

Before normalization, we check for **outlier samples** — samples whose global intensity profile is so different from the rest that they would distort the normalization and downstream statistics. We use a **4-method consensus approach**: a sample is flagged only if at least 3 of 4 independent methods agree it is an outlier [@Huang2024].

This consensus strategy provides high **specificity** (low false-positive rate). A single method might flag a sample due to an idiosyncratic metric, but agreement across multiple independent metrics provides strong evidence.

### Method 1: Sample missingness

We flag samples with unusually high missingness or a large difference in missingness between their Pre and Post timepoints.

```{r}
#| label: outlier-missingness

pct_missing <- colMeans(is.na(dal$data)) * 100

miss_info <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(pct_missing = pct_missing[Col_ID],
         prefix = str_remove(Col_ID, "_(Pre|Post)$")) |>
  group_by(prefix) |>
  mutate(delta_missing = abs(diff(pct_missing))) |>
  ungroup()

# IQR-based thresholds (standard Tukey fence)
miss_thresh  <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing)
delta_thresh <- quantile(miss_info$delta_missing, 0.75) +
  1.5 * IQR(miss_info$delta_missing)

miss_info$miss_flag <- miss_info$pct_missing > miss_thresh |
  miss_info$delta_missing > delta_thresh
```

### Method 2: PCA Mahalanobis distance

We project all samples into PCA space (using the first 3 principal components) and compute the **Mahalanobis distance** from each sample to the multivariate centroid. Samples far from the center — beyond a chi-squared cutoff at p < `r cfg$mahal_p` — are flagged.

```{r}
#| label: outlier-pca

complete_mat <- dal$data[rowSums(is.na(dal$data)) == 0, ]
pca_pre  <- run_pca(complete_mat, dal$metadata, log_transform = TRUE)
pc3      <- pca_pre$pca$x[, 1:3]
mahal    <- mahalanobis(pc3, colMeans(pc3), cov(pc3))
pca_flags <- tibble(Col_ID = colnames(dal$data), mahal_dist = mahal,
                    pca_flag = mahal > qchisq(1 - cfg$mahal_p, df = 3))
```

### Method 3: MAD-based median intensity

We compute the median log2 intensity for each sample. Samples whose median deviates by more than `r cfg$mad_k` x MAD (Median Absolute Deviation) from the global median are flagged. **MAD** is a robust measure of spread — unlike standard deviation, it is not inflated by the very outliers we are trying to detect.

```{r}
#| label: outlier-mad

samp_med   <- apply(log2(dal$data + 1), 2, median, na.rm = TRUE)
global_med <- median(samp_med)
mad_val    <- mad(samp_med)
mad_flags  <- tibble(Col_ID = names(samp_med), sample_median = samp_med,
                     mad_flag = abs(samp_med - global_med) > cfg$mad_k * mad_val)
```

### Method 4: Inter-sample correlation

We compute pairwise Pearson correlations between all samples (on log2 intensities) and take each sample's median correlation with all other samples. Samples with unusually low median correlation are flagged.

```{r}
#| label: outlier-correlation

cor_mat     <- cor(log2(dal$data + 1), use = "pairwise.complete.obs")
med_cor     <- apply(cor_mat, 2, function(x) median(x[x < 1], na.rm = TRUE))
cor_med_all <- median(med_cor)
cor_mad_all <- mad(med_cor)
cor_flags   <- tibble(Col_ID = names(med_cor), median_cor = med_cor,
                      cor_flag = med_cor < cor_med_all - cfg$mad_k * cor_mad_all)
```

### Consensus decision

A sample is removed if **at least 3 of 4 methods** flag it:

```{r}
#| label: outlier-consensus

outlier_diag <- miss_info |>
  left_join(pca_flags, by = "Col_ID") |>
  left_join(mad_flags, by = "Col_ID") |>
  left_join(cor_flags, by = "Col_ID") |>
  mutate(n_flags = miss_flag + pca_flag + mad_flag + cor_flag,
         consensus_outlier = n_flags >= cfg$outlier_k)

n_outliers <- sum(outlier_diag$consensus_outlier)
cat(sprintf("Outliers: %d sample(s) flagged (%d/4 consensus)\n",
            n_outliers, cfg$outlier_k))

# Snapshot pre-outlier state for diagnostics
data_pre_outlier <- dal$data
meta_pre_outlier <- dal$metadata

# Remove outlier samples
outlier_ids <- outlier_diag |> filter(consensus_outlier) |> pull(Col_ID)
if (n_outliers > 0) {
  dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
  cat(sprintf("Removed: %s (%d remain)\n",
              paste(outlier_ids, collapse = ", "), ncol(dal$data)))
}
```

Two samples (Y_S05_Pre and Y_S05_Post) were flagged by the consensus. Both timepoints of this subject are removed to maintain the **paired design** — keeping one timepoint without the other would break the repeated-measures correlation structure used in the DEP phase.

After outlier removal, we have `r ncol(dal$data)` samples: 16 Young (Pre + Post) and 15 Old (Pre + Post).

---

## Step 6: Normalization

**Normalization** corrects for systematic technical biases between samples — differences in total protein loaded, instrument sensitivity drift, or other run-to-run variation — so that remaining intensity differences reflect genuine biology rather than technical artifacts.

We use **cyclic loess** (cycloess) normalization [@Bolstad2003], implemented in proteoDA. Cyclic loess works by iteratively applying loess regression between all pairs of samples to remove intensity-dependent biases. It was chosen because:

- It handles **intensity-dependent biases** (where the magnitude of the bias depends on the protein's abundance), unlike simpler methods like median centering
- It does not assume symmetric fold-change distributions, unlike quantile normalization
- It was empirically validated on this dataset using `proteoDA::write_norm_report()`, which compares multiple normalization methods side-by-side [@Arend2025]
- It is well-established for label-free proteomics [@Valikangas2018]

### Generate comparison report and apply normalization

```{r}
#| label: normalize

# Generate multi-method comparison report
write_norm_report(dal, grouping_column = "Group_Time",
                  output_dir = cfg$report_dir,
                  filename = "01_norm_comparison.pdf", overwrite = TRUE)

# Pre-normalization QC report
write_qc_report(dal, color_column = "Group_Time",
                output_dir = cfg$report_dir,
                filename = "02_qc_pre.pdf", overwrite = TRUE)

# Save pre-normalization snapshot
dal_pre <- dal
saveRDS(dal_pre, file.path(cfg$data_dir, "01_DAList_prenorm.rds"))

# Apply cyclic loess normalization
dal <- normalize_data(dal, norm_method = cfg$norm_method)
cat(sprintf("Normalized (%s): %d proteins x %d samples\n",
            cfg$norm_method, nrow(dal$data), ncol(dal$data)))

# Post-normalization QC report
write_qc_report(dal, color_column = "Group_Time",
                output_dir = cfg$report_dir,
                filename = "03_qc_post.pdf", overwrite = TRUE)
```

The `write_norm_report()` function generates `01_norm_comparison.pdf`, which shows box plots, PCA, and CV metrics for multiple normalization methods on this specific dataset. This empirical comparison supports the choice of cycloess. The pre- and post-normalization QC reports (`02_qc_pre.pdf` and `03_qc_post.pdf`) let you visually confirm that normalization improved sample alignment without distorting biological group structure.

> **Note:** Cycloess assumes that **most proteins are unchanged** between conditions. This is a standard and reasonable assumption for exercise proteomics, where training-induced changes are typically bidirectional and affect a minority of proteins. If the intervention caused widespread unidirectional changes (all proteins up or all down), cycloess would partially "normalize away" this global shift.

---

## Step 7: Export Normalized Data

We export two key files that will be consumed by downstream stages:

```{r}
#| label: export

export_df <- bind_cols(
  as_tibble(dal$annotation) |> select(uniprot_id, protein, gene, description),
  as_tibble(dal$data))

write_csv(export_df, file.path(cfg$data_dir, "02_normalized.csv"))
saveRDS(dal, file.path(cfg$data_dir, "03_DAList_normalized.rds"))
```

| File | Description | Used by |
|------|-------------|---------|
| `02_normalized.csv` | Normalized matrix (4 annotation cols + 62 sample cols) | Imputation (Stage 02), DEP (Stage 03), all figure scripts |
| `03_DAList_normalized.rds` | Full DAList R object with metadata | DEP (Stage 03) for metadata extraction |

The CSV file is the **primary handoff artifact** — it is the input for both imputation (Phase 2) and limma differential expression (Phase 3, which runs on non-imputed data).

---

## Step 8: Compute Diagnostics and Save Report Intermediates

The final section computes diagnostic data (filter summaries, outlier details, variability metrics) and saves them to an intermediates file. The companion script `02_norm_reports.R` reads this file to generate a 4-page diagnostic PDF and a supplementary Excel workbook — keeping computation and visualization separate.

```{r}
#| label: diagnostics

# Prepare filter bar chart data
filter_bar_data <- filter_log |>
  filter(!is.na(n_removed)) |>
  mutate(step = factor(step, levels = step)) |>
  pivot_longer(c(n_after, n_removed), names_to = "status", values_to = "n") |>
  mutate(status = recode(status, n_after = "Retained", n_removed = "Removed"))

# Prepare missingness bar data (all samples, including outliers)
miss_bar_data <- meta_pre_outlier |>
  select(Col_ID, Group_Time) |>
  mutate(detected = colSums(!is.na(data_pre_outlier[, Col_ID])),
         missing  = nrow(data_pre_outlier) - detected,
         is_outlier = Col_ID %in% outlier_ids) |>
  pivot_longer(c(detected, missing), names_to = "status", values_to = "n") |>
  mutate(status = str_to_title(status))

# Per-subject variability (IQR of log2 intensities)
subj_var <- dal$metadata |>
  mutate(iqr = apply(log2(dal$data[, Col_ID] + 1), 2, IQR, na.rm = TRUE),
         Subject_ID = str_remove(Col_ID, "_(Pre|Post)$")) |>
  select(Col_ID, Subject_ID, Group, Timepoint, Group_Time, iqr)

# Eta-squared: proportion of variance explained by group membership
log_dat <- log2(dal$data + 1)
grp_vec <- dal$metadata$Group_Time[match(colnames(log_dat), dal$metadata$Col_ID)]
eta2_vals <- apply(log_dat, 1, function(x) {
  ok <- !is.na(x)
  if (sum(ok) < 4) return(NA_real_)
  xk <- x[ok]; gk <- grp_vec[ok]
  ss_b <- sum(tapply(xk, gk, length) * (tapply(xk, gk, mean) - mean(xk))^2)
  ss_t <- sum((xk - mean(xk))^2)
  if (ss_t > 0) ss_b / ss_t else NA_real_
})

# Post-normalization PCA
pca_post <- run_pca(dal$data, dal$metadata, log_transform = FALSE)

# Save all intermediates for the report script
intermediates <- list(
  cfg              = cfg,
  filter_log       = filter_log,
  filter_bar_data  = filter_bar_data,
  miss_bar_data    = miss_bar_data,
  n_raw            = n_raw,
  n_outliers       = n_outliers,
  outlier_diag     = outlier_diag,
  outlier_ids      = outlier_ids,
  miss_thresh      = miss_thresh,
  delta_thresh     = delta_thresh,
  pca_pre          = pca_pre,
  pca_post         = pca_post,
  global_med       = global_med,
  mad_val          = mad_val,
  subj_var         = subj_var,
  eta2_vals        = eta2_vals,
  filtered_proteins = filtered_proteins,
  data_pre_outlier = data_pre_outlier,
  meta_pre_outlier = meta_pre_outlier,
  dal_nrow         = nrow(dal$data),
  dal_ncol         = ncol(dal$data)
)

saveRDS(intermediates, file.path(cfg$data_dir, "00_report_intermediates.rds"))

cat(sprintf("Done: %d proteins x %d samples -> %s/\n",
            nrow(dal$data), ncol(dal$data), cfg$data_dir))
```

The **eta-squared** ($\eta^2$) metric tells us what fraction of each protein's total variance is explained by group membership (Young_Pre, Young_Post, Old_Pre, Old_Post). A high median $\eta^2$ indicates that between-group differences dominate over within-group noise, which is good news for downstream differential expression analysis.

---

## Output Summary

### Data outputs (`c_data/`)

| File | Description |
|------|-------------|
| `00_report_intermediates.rds` | All diagnostic data for `02_norm_reports.R` |
| `01_DAList_prenorm.rds` | DAList snapshot before normalization |
| `02_normalized.csv` | **Canonical output**: normalized matrix (2,124 x 66: 4 annot + 62 samples) |
| `03_DAList_normalized.rds` | **Canonical output**: full DAList with metadata |
| `05_normalization_supp.xlsx` | Supplementary workbook (generated by `02_norm_reports.R`) |

### Report outputs (`b_reports/`)

| Report | Contents |
|--------|----------|
| `01_norm_comparison.pdf` | Multi-method normalization comparison (box plots, PCA, CV metrics) |
| `02_qc_pre.pdf` | Pre-normalization QC (distributions, PCA, clustering) |
| `03_qc_post.pdf` | Post-normalization QC (same metrics — should show improvement) |
| `04_diagnostics.pdf` | 4-page diagnostic report (generated by `02_norm_reports.R`) |

---

## Results and Interpretation

### Protein retention

```{r}
#| label: filter-summary

knitr::kable(filter_log, caption = "Protein retention through filtering pipeline",
             digits = 1)
```

Starting from `r n_raw` raw proteins, `r round(nrow(dal$data) / n_raw * 100, 1)`% (`r nrow(dal$data)`) are retained after all filtering steps. The largest loss comes from missingness filtering (`r filter_log$n_removed[filter_log$step == sprintf("Missingness (>=%d%% in >=1 group)", cfg$miss_pct * 100)]` proteins), which removes proteins that were only sporadically detected. This retention rate is typical for DIA-MS skeletal muscle proteomics.

### Outlier detection

Only `r n_outliers` subject(s) were flagged across `r cfg$outlier_k`/4 methods, indicating a genuinely aberrant sample rather than a borderline case. Removing both Pre and Post timepoints maintains the paired design required for `duplicateCorrelation` in the DEP phase.

### Final dataset

The normalized dataset contains **`r nrow(dal$data)` proteins x `r ncol(dal$data)` samples**:

| Group | Pre | Post | Subjects |
|-------|-----|------|----------|
| Young | 16 | 16 | 16 |
| Old | 15 | 15 | 15 |

Missing values remain in the normalized matrix (`r sum(is.na(dal$data))` cells, `r round(sum(is.na(dal$data)) / length(dal$data) * 100, 1)`% of all values). These will be addressed in Phase 2 (Imputation).

---

## Validation and Limitations

### What to look for in the QC reports

- **PCA (post-norm):** Samples should partially separate by Age (PC1 or PC2), with Pre and Post overlapping within each age group. Complete separation by timepoint would be unexpected — age effects are larger than exercise effects in skeletal muscle proteomics.
- **Box plots (post-norm):** Sample-level intensity distributions should be well-aligned after cycloess, with no systematic shifts between groups.
- **Diagnostics page 2 (outliers):** Flagged samples should be clear outliers in at least 3 of the 4 panels.

### Known limitations

1. **HPA filter may be conservative** — Some biologically relevant proteins may lack HPA skeletal muscle annotation. Check the `Filtered_Proteins` sheet in the supplementary workbook.
2. **Cycloess assumes most proteins unchanged** — Reasonable for exercise proteomics (changes are bidirectional and affect a minority of proteins), but would be problematic for interventions causing global unidirectional shifts.
3. **Outlier removal is irreversible** — The 3/4 consensus threshold is conservative, but still removes `r length(outlier_ids)` samples. The DEP reporting script includes an outlier-removal sensitivity analysis comparing results with and without outlier removal.

Now that we have a normalized expression matrix with `r nrow(dal$data)` proteins across `r ncol(dal$data)` samples, we move to characterizing and handling the remaining `r round(sum(is.na(dal$data)) / length(dal$data) * 100, 1)`% missing values in Phase 2 (Imputation).

---

## Downstream Dependencies

The normalized DAList (`03_DAList_normalized.rds`) is the primary input for:

- **Phase 2 (Imputation):** Uses the normalized matrix and missingness patterns
- **Phase 3 (DEP):** Uses the normalized matrix directly for limma analysis (non-imputed)
- **All figure scripts:** Load metadata from this DAList

> **Note:** The DAList stores `Subject_ID` in its metadata, but `Subject_ID` is **not globally unique** — 8 IDs are shared between the Young and Old cohorts (different individuals). Downstream scripts that need subject-level pairing (e.g., `duplicateCorrelation`) derive a unique key from `Col_ID`: `sub("_(Pre|Post)$", "", Col_ID)`.

---

## References

::: {#refs}
:::

---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
