---
title: "YvO Proteomics Pipeline — Phase 2: Imputation"
author: "DTL"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
bibliography: ../../references.bib
---

# Overview

In Phase 1 (Normalization), we produced a clean, normalized expression matrix of **2,124 proteins x 62 samples** — but roughly 11.7% of the values are still missing. In this tutorial, we address those missing values through **imputation**: the statistical estimation of what the missing values most likely were.

Why not just ignore missing values? Two reasons:

1. **Downstream methods need complete data.** Clustering (Phase 4), co-expression networks (Phase 5), and some visualization methods cannot handle `NA`s.
2. **Removing incomplete proteins would be devastating.** Complete-case analysis (keeping only proteins with zero missing values) would discard over half of our 2,124 proteins [@WebbRobertson2015], including many biologically interesting low-abundance proteins.

The key insight driving our approach is that **not all missing values arise from the same mechanism**. We classify each protein's missingness as either **Missing At Random (MAR)** or **Missing Not At Random (MNAR)**, then benchmark 12 imputation methods — including hybrid methods that handle each type differently — to find the best performer for this dataset.

> **Note:** The primary differential expression analysis (Phase 3) runs on **non-imputed** data. Limma handles missing values per-protein, and running DEP on non-imputed data avoids introducing imputation artifacts into our statistical tests [@Karpievitch2012]. The imputed matrix produced here is used for visualization, clustering, and as a sensitivity check.

**Input:** `01_normalization/c_data/02_normalized.csv` (2,124 proteins x 62 samples, ~11.7% missing)

**Output:** `02_Imputation/c_data/01_imputed.csv` (2,124 proteins x 62 samples, 0% missing)

**Scripts:** `01_run_imputation.R` (core computation) and `02_imputation_reports.R` (visualization)

---

## Prerequisites

### Required packages

```{r}
#| label: setup

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(MsCoreUtils)
library(msImpute)
library(pcaMethods)
library(imputeLCMD)
library(missForest)
library(missMDA)
library(vegan)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(stringr)
```

Key packages and why we use them:

- **MsCoreUtils** — Provides `impute_matrix()`, a unified interface for multiple imputation methods (MinProb, MinDet, QRILC, knn, bpca, mixed). It wraps methods from several packages into a consistent API.
- **msImpute** [@HediyehZadeh2023] — Provides `selectFeatures()` with the entropy-based method (EBM) for classifying proteins as MAR vs MNAR. This is our primary classification approach.
- **pcaMethods** — SVD-based imputation from Bioconductor.
- **imputeLCMD** — Left-censored imputation methods (QRILC) designed specifically for MNAR data.
- **missForest** — Random Forest imputation (a non-parametric, structure-preserving method).
- **missMDA** — Regularized PCA imputation.
- **vegan** — Provides `procrustes()` for measuring how well imputation preserves the global PCA geometry.

### Configuration

```{r}
#| label: config

set.seed(42)

cfg <- list(
  NORM_CSV        = "01_normalization/c_data/02_normalized.csv",
  NORM_RDS        = "01_normalization/c_data/03_DAList_normalized.rds",
  DATA_DIR        = "02_Imputation/c_data",
  N_ITER          = 20L,        # benchmark iterations
  MASK_FRAC       = 0.10,       # fraction of observed values to mask per iteration
  MISS_UNRELIABLE = 50,         # proteins with >50% missing flagged as unreliable
  METHODS = list(
    # MNAR (left-censored) methods
    MinProb        = list(method = "MinProb"),
    MinDet         = list(method = "MinDet"),
    QRILC          = list(method = "QRILC"),
    zero           = list(method = "zero"),
    # MAR (structure-based) methods
    knn            = list(method = "knn"),
    bpca           = list(method = "bpca"),
    RF             = list(method = "RF"),
    SVD            = list(method = "SVD"),
    imputePCA      = list(method = "imputePCA"),
    # Hybrid (MAR backbone + QRILC for MNAR)
    mix_bpca_QRILC = list(method = "mixed", mar = "bpca", mnar = "QRILC"),
    mix_knn_QRILC  = list(method = "mixed", mar = "knn",  mnar = "QRILC"),
    mix_RF_QRILC   = list(method = "mixed", mar = "RF",   mnar = "QRILC")
  )
)

METHOD_TYPE <- tibble(method = names(cfg$METHODS)) |>
  mutate(type = case_when(
    method %in% c("MinProb", "MinDet", "QRILC", "zero") ~ "MNAR",
    str_starts(method, "mix_") ~ "Hybrid",
    TRUE ~ "MAR"))

dir.create(cfg$DATA_DIR, showWarnings = FALSE, recursive = TRUE)
```

### Helper functions

```{r}
#| label: helpers

# Unified imputation wrapper
run_impute <- function(m, mat, randna) {
  if (m$method == "mixed")
    MsCoreUtils::impute_matrix(mat, method = "mixed", randna = randna,
                               mar = m$mar, mnar = m$mnar)
  else if (m$method == "imputePCA")
    missMDA::imputePCA(mat, ncp = 2, method = "Regularized")$completeObs
  else if (m$method == "SVD")
    pcaMethods::pca(mat, method = "svdImpute", nPcs = 2, verbose = FALSE)@completeObs
  else
    MsCoreUtils::impute_matrix(mat, method = m$method)
}

# Normalized Root Mean Squared Error
nrmse <- function(true_vals, imp_vals) {
  sqrt(mean((true_vals - imp_vals)^2)) / sd(true_vals)
}

# Procrustes Sum of Squares: PCA structure preservation (lower = less distortion)
pss_metric <- function(original, imputed, n_pc = 5) {
  complete_prots <- complete.cases(original)
  if (sum(complete_prots) < n_pc + 1) return(NA_real_)
  n_pc_use <- min(n_pc, ncol(original) - 1)
  pca_orig <- prcomp(t(original[complete_prots, ]), center = TRUE, scale. = TRUE)
  pca_imp  <- prcomp(t(imputed[complete_prots, ]),  center = TRUE, scale. = TRUE)
  vegan::procrustes(pca_orig$x[, seq_len(n_pc_use)],
                    pca_imp$x[, seq_len(n_pc_use)])$ss
}
```

The **NRMSE** (Normalized Root Mean Squared Error) measures point-wise accuracy: how close are the imputed values to the known truth? Lower is better.

The **Procrustes Sum of Squares** (PSS) measures whether imputation distorts the global sample relationships. It compares PCA projections of the original and imputed matrices by finding the best rotation/scaling alignment and measuring the residual discrepancy. A method could have low NRMSE (accurate per-value) but high PSS (distorts the overall structure).

---

## Step 1: Load Normalized Data

```{r}
#| label: load-data

df  <- read_csv(cfg$NORM_CSV, show_col_types = FALSE)
ann <- df |> select(uniprot_id, gene, protein, description)
mat <- as.matrix(df[, -(1:4)])  # everything after the 4 annotation columns
rownames(mat) <- df$gene

# Handle rare case of duplicate gene names
if (any(duplicated(df$gene))) {
  warning("Duplicate gene names; using uniprot_id as rownames")
  rownames(mat) <- df$uniprot_id
}

cat(sprintf("%d proteins x %d samples\n", nrow(mat), ncol(mat)))

# Metadata from normalized DAList (not regex-derived)
dal_norm <- readRDS(cfg$NORM_RDS)
meta <- tibble(
  Col_ID     = dal_norm$metadata$Col_ID,
  Group      = dal_norm$metadata$Group,
  Timepoint  = dal_norm$metadata$Timepoint,
  Group_Time = dal_norm$metadata$Group_Time
)
stopifnot(setequal(meta$Col_ID, colnames(mat)))
```

---

## Step 2: Characterize Missingness

Before imputing, we need to understand the missingness landscape. How many proteins have missing values? How severe is the missingness? Is it concentrated in certain groups?

```{r}
#| label: missingness-profile

prot_miss <- rowSums(is.na(mat))
prot_pct  <- prot_miss / ncol(mat) * 100
obs_means <- rowMeans(mat, na.rm = TRUE)
pct_miss  <- round(sum(is.na(mat)) / length(mat) * 100, 2)

cat(sprintf("Missing: %d / %d (%.2f%%) | Complete: %d proteins\n",
            sum(is.na(mat)), length(mat), pct_miss, sum(prot_miss == 0)))

# Per-group missingness
miss_by_group <- sapply(unique(meta$Group_Time), function(g) {
  cols <- meta$Col_ID[meta$Group_Time == g]
  rowSums(is.na(mat[, cols, drop = FALSE])) / length(cols) * 100
})
```

About `r pct_miss`% of all values are missing. Roughly half of proteins have zero missing values (complete cases), while the rest have varying degrees of missingness.

---

## Step 3: MAR/MNAR Classification

This is the conceptual heart of the imputation strategy. We classify each protein with missing values as either:

- **MAR (Missing At Random):** The missingness is unrelated to the protein's true abundance. It might be caused by peptide interference, chromatographic issues, or random detection failure. MAR proteins can be imputed using structure-preserving methods (knn, bpca, etc.) that leverage patterns across samples.

- **MNAR (Missing Not At Random):** The protein is missing *because* its abundance is below the instrument's detection limit. Low-abundance proteins are systematically more likely to be missing. MNAR proteins need left-censored imputation methods (MinProb, QRILC) that draw values from the lower tail of the intensity distribution.

We use **msImpute EBM** [@HediyehZadeh2023] as our primary classifier. This entropy-based method examines the information content of each protein's missingness pattern to determine whether it is consistent with MAR or MNAR.

```{r}
#| label: classification

has_na <- which(prot_miss > 0 & prot_miss < ncol(mat))
miss_class <- tibble(gene = rownames(mat), n_miss = prot_miss,
                     pct_miss = prot_pct, mean_intensity = obs_means)

# Primary: msImpute EBM | Fallback: k-means on (intensity, missingness)
mar_result <- tryCatch({
  feat <- msImpute::selectFeatures(mat[has_na, ], method = "ebm",
                                   group = meta$Group_Time)
  mar_names <- feat$name[feat$msImpute_feature == TRUE]
  n_incomplete <- nrow(feat)
  cat(sprintf("EBM result: %d MAR / %d MNAR (of %d incomplete proteins)\n",
              length(mar_names), n_incomplete - length(mar_names), n_incomplete))
  # Guard against degenerate classification
  if (length(mar_names) < 0.05 * n_incomplete) {
    cat("  EBM degenerate (<5% MAR) -- falling back to k-means\n")
    NULL
  } else {
    list(mar_genes = mar_names, method = "msImpute_ebm")
  }
}, error = function(e) {
  cat(sprintf("msImpute EBM failed (%s) -- falling back to k-means\n",
              conditionMessage(e)))
  NULL
})

# Fallback: k-means clustering on standardized (intensity, missingness)
if (is.null(mar_result)) {
  mc_sub <- miss_class |> filter(n_miss > 0, n_miss < ncol(mat))
  km <- kmeans(scale(cbind(mc_sub$mean_intensity, mc_sub$pct_miss)),
               centers = 2, nstart = 25)
  cl_means <- tapply(mc_sub$mean_intensity, km$cluster, mean)
  mnar_cl <- which.min(cl_means)  # low-intensity cluster = MNAR
  mar_result <- list(
    mar_genes = mc_sub$gene[km$cluster != mnar_cl],
    method = sprintf("k-means (cluster means: %.1f vs %.1f)",
                     cl_means[mnar_cl], cl_means[-mnar_cl]))
}

cat(sprintf("Classification: %s\n", mar_result$method))

miss_class <- miss_class |>
  mutate(
    classification = case_when(
      n_miss == 0 ~ "Complete",
      gene %in% mar_result$mar_genes ~ "MAR",
      TRUE ~ "MNAR"),
    imputation_reliable = classification == "Complete" | pct_miss < cfg$MISS_UNRELIABLE
  )
```

The `imputation_reliable` flag follows @Webel2024, who showed that downstream differential expression performance degrades substantially when proteins exceed ~50% missingness — imputed values begin to dominate the estimated means. Proteins flagged as unreliable are still imputed and retained in the matrix, but should be interpreted with extra caution.

### Group-stratified missingness test

For each MNAR protein, we run a Fisher exact test to check whether its missingness is uniformly distributed across the four experimental groups. Proteins with significantly non-uniform missingness might have their missingness confounded with biological signal.

```{r}
#| label: group-missingness

mnar_genes <- miss_class$gene[miss_class$classification == "MNAR"]
group_miss_pval <- setNames(rep(NA_real_, nrow(miss_class)), miss_class$gene)

for (g in mnar_genes) {
  ct <- sapply(unique(meta$Group_Time), function(gt) {
    cols <- meta$Col_ID[meta$Group_Time == gt]
    c(missing = sum(is.na(mat[g, cols])), observed = sum(!is.na(mat[g, cols])))
  })
  group_miss_pval[g] <- tryCatch(
    fisher.test(ct, simulate.p.value = TRUE, B = 2000)$p.value,
    error = function(e) NA_real_)
}
miss_class$group_miss_pval <- group_miss_pval[miss_class$gene]

n_sig <- sum(group_miss_pval[mnar_genes] < 0.05, na.rm = TRUE)

# Summary at both protein and missing-value levels
n_mar_prots  <- sum(miss_class$classification == "MAR")
n_mnar_prots <- length(mnar_genes)
n_comp_prots <- sum(miss_class$classification == "Complete")
mar_miss_vals  <- sum(miss_class$n_miss[miss_class$classification == "MAR"])
mnar_miss_vals <- sum(miss_class$n_miss[miss_class$classification == "MNAR"])
total_miss_vals <- mar_miss_vals + mnar_miss_vals

cat(sprintf("Proteins: MAR %d | MNAR %d (%d group-biased) | Complete %d\n",
            n_mar_prots, n_mnar_prots, n_sig, n_comp_prots))
cat(sprintf("Values: MAR %d (%.0f%%) | MNAR %d (%.0f%%)\n",
            mar_miss_vals, mar_miss_vals / total_miss_vals * 100,
            mnar_miss_vals, mnar_miss_vals / total_miss_vals * 100))
```

The classification scatter plot (in the missingness report) shows the result visually: MAR proteins (blue) are scattered across the intensity range with moderate missingness, while MNAR proteins (red) cluster at low intensity with high missingness.

---

## Step 4: Benchmark 12 Imputation Methods

We benchmark all 12 methods using a **masking-based** approach [@Jin2021]:

1. For each of `r cfg$N_ITER` iterations, randomly mask `r cfg$MASK_FRAC * 100`% of observed values from MAR/Complete proteins
2. Run all 12 methods on the masked matrix
3. Compare imputed values to the known truth using NRMSE and PSS

Only MAR protein values are masked, because we have no ground truth for MNAR values — by definition, they were never reliably observed.

```{r}
#| label: benchmark

# randna: TRUE = MAR/Complete (apply MAR method), FALSE = MNAR (apply MNAR method)
randna <- setNames(miss_class$classification != "MNAR", miss_class$gene)

# Pre-compute intensity bins for per-intensity analysis
bench_genes <- miss_class$gene[randna[miss_class$gene]]
bench_means <- rowMeans(mat[bench_genes, , drop = FALSE], na.rm = TRUE)
bin_breaks  <- unique(quantile(bench_means, probs = c(0, 1/3, 2/3, 1)))
bin_labels  <- c("low", "mid", "high")[seq_len(length(bin_breaks) - 1)]
prot_bins   <- cut(bench_means, breaks = bin_breaks, labels = bin_labels,
                   include.lowest = TRUE)
names(prot_bins) <- bench_genes

cat(sprintf("Benchmarking: %d methods x %d iterations\n",
            length(cfg$METHODS), cfg$N_ITER))
bench_raw <- vector("list", length(cfg$METHODS) * cfg$N_ITER)
bin_raw   <- vector("list", length(cfg$METHODS) * cfg$N_ITER * length(bin_labels))
k <- 0L; kb <- 0L

for (iter in seq_len(cfg$N_ITER)) {
  if (iter %% 5 == 0) cat(sprintf("  Iter %d/%d\n", iter, cfg$N_ITER))

  # Mask MASK_FRAC of observed values from MAR/Complete proteins
  mar_obs_idx <- which(!is.na(mat) & randna[row(mat)])
  mask_idx    <- sample(mar_obs_idx, round(length(mar_obs_idx) * cfg$MASK_FRAC))
  true_v      <- mat[mask_idx]
  mm <- mat; mm[mask_idx] <- NA

  # Row indices for per-intensity binning
  mask_rows <- rownames(mat)[((mask_idx - 1) %% nrow(mat)) + 1]
  mask_bin  <- prot_bins[mask_rows]

  for (nm in names(cfg$METHODS)) {
    imp <- tryCatch(run_impute(cfg$METHODS[[nm]], mm, randna),
                    error = function(e) NULL)
    if (is.null(imp)) next

    k <- k + 1L
    nrmse_val <- nrmse(true_v, imp[mask_idx])
    pss_val   <- tryCatch(pss_metric(mat, imp), error = function(e) NA_real_)
    bench_raw[[k]] <- tibble(method = nm, iter = iter,
                             nrmse = nrmse_val, pss = pss_val)

    # Per-intensity bin NRMSE
    for (b in bin_labels) {
      sel <- which(mask_bin == b & !is.na(mask_bin))
      if (length(sel) < 5) next
      kb <- kb + 1L
      bin_raw[[kb]] <- tibble(method = nm, iter = iter, bin = b,
                              nrmse = nrmse(true_v[sel], imp[mask_idx[sel]]))
    }
  }
}

bench_df  <- bind_rows(bench_raw)
bench_sum <- bench_df |>
  group_by(method) |>
  summarise(mean_nrmse   = mean(nrmse),
            sd_nrmse     = sd(nrmse),
            median_nrmse = median(nrmse),
            mean_pss     = mean(pss, na.rm = TRUE),
            sd_pss       = sd(pss, na.rm = TRUE),
            .groups = "drop") |>
  arrange(mean_nrmse)

bin_df  <- bind_rows(bin_raw)
bin_sum <- bin_df |>
  group_by(method, bin) |>
  summarise(mean_nrmse = mean(nrmse), sd_nrmse = sd(nrmse), .groups = "drop")

# Extended summary for top 5
top5 <- bench_sum$method[1:min(5, nrow(bench_sum))]
ext_sum <- bench_sum |>
  filter(method %in% top5) |>
  select(method, nrmse = mean_nrmse, pss = mean_pss) |>
  left_join(
    bin_sum |> filter(method %in% top5) |>
      select(method, bin, mean_nrmse) |>
      pivot_wider(names_from = bin, values_from = mean_nrmse, names_prefix = "nrmse_"),
    by = "method")

best <- bench_sum$method[1]
cat(sprintf("Best: %s (NRMSE %.4f, PSS %.4f)\n",
            best, bench_sum$mean_nrmse[1], bench_sum$mean_pss[1]))
```

### Interpreting the benchmark results

```{r}
#| label: benchmark-results

knitr::kable(bench_sum[1:5, ], caption = "Top 5 methods by NRMSE", digits = 4)
```

Hybrid methods (mixing a MAR method with an MNAR method) often outperform single-method approaches [@Lazar2016; @Borcherding2021]. However, at low overall missingness (~12%), pure MAR methods may match or beat hybrids because even MNAR-classified proteins retain enough observed values for structure-preserving algorithms like bpca to fit reliable models [@HediyehZadeh2023].

> **Note:** If the benchmark winner is a pure MAR method (e.g., bpca), the MAR/MNAR classification does not affect the imputed values — all proteins receive the same treatment. The classification is still useful for reporting and the MNAR audit.

---

## Step 5: Apply the Best Method

We apply the winning method to the full matrix (with its original missing values, not the artificial masks):

```{r}
#| label: apply-best

set.seed(42)
mat_imp <- run_impute(cfg$METHODS[[best]], mat, randna)
stopifnot(sum(is.na(mat_imp)) == 0)  # verify no NAs remain
cat(sprintf("Imputation complete: %d proteins x %d samples, 0 missing\n",
            nrow(mat_imp), ncol(mat_imp)))
```

The `randna` vector tells hybrid methods which proteins are MAR vs MNAR, so they can apply the appropriate sub-method to each.

---

## Step 6: MNAR Audit

For each MNAR protein, we compare the mean intensity before and after imputation. Since MNAR methods impute values at the lower tail of the distribution, we expect post-imputation means to be lower than pre-imputation means (negative shift).

```{r}
#| label: mnar-audit

mnar_audit <- tibble(
  gene      = mnar_genes,
  pre_mean  = rowMeans(mat[mnar_genes, ], na.rm = TRUE),
  post_mean = rowMeans(mat_imp[mnar_genes, ]),
  pre_sd    = apply(mat[mnar_genes, ], 1, sd, na.rm = TRUE),
  pct_miss  = prot_pct[mnar_genes],
  shift     = rowMeans(mat_imp[mnar_genes, ]) -
              rowMeans(mat[mnar_genes, ], na.rm = TRUE),
  effect_d  = (rowMeans(mat_imp[mnar_genes, ]) -
               rowMeans(mat[mnar_genes, ], na.rm = TRUE)) /
              apply(mat[mnar_genes, ], 1, sd, na.rm = TRUE),
  imputation_reliable = prot_pct[mnar_genes] < cfg$MISS_UNRELIABLE
)

cat(sprintf("MNAR audit: %d proteins | Median shift: %.3f log2 units\n",
            nrow(mnar_audit), median(mnar_audit$shift)))
cat(sprintf("Reliable (<50%% missing): %d (%.1f%%)\n",
            sum(mnar_audit$imputation_reliable),
            100 * mean(mnar_audit$imputation_reliable)))
```

A negative median shift is expected and confirms that MNAR imputation is working as intended — values are being drawn from the low-intensity tail. Large *positive* shifts would indicate problematic imputation.

---

## Step 7: Export Results

```{r}
#| label: export

was_na <- is.na(mat)
stopifnot(identical(ann$gene, rownames(mat_imp)))

# Imputed matrix
write_csv(bind_cols(ann, as_tibble(mat_imp)),
          file.path(cfg$DATA_DIR, "01_imputed.csv"))

# Updated DAList with missingness annotations
dal <- readRDS(cfg$NORM_RDS)
mat_imp_uid <- mat_imp
rownames(mat_imp_uid) <- ann$uniprot_id
dal$data <- mat_imp_uid
dal$annotation <- merge(
  dal$annotation,
  miss_class |> select(gene, n_miss, pct_miss,
                       miss_classification = classification,
                       imputation_reliable),
  by = "gene", all.x = TRUE, sort = FALSE)
saveRDS(dal, file.path(cfg$DATA_DIR, "01_DAList_imputed.rds"))

# Supporting tables
write_csv(miss_class, file.path(cfg$DATA_DIR, "02_mar_mnar_classification.csv"))
write_csv(bench_sum,  file.path(cfg$DATA_DIR, "03_benchmark_summary.csv"))
write_csv(bench_df,   file.path(cfg$DATA_DIR, "04_benchmark_raw_iterations.csv"))
write_csv(ext_sum,    file.path(cfg$DATA_DIR, "05_benchmark_extended.csv"))
write_csv(bin_sum,    file.path(cfg$DATA_DIR, "06_benchmark_per_intensity.csv"))
write_csv(bind_cols(tibble(gene = rownames(was_na)), as_tibble(was_na)),
          file.path(cfg$DATA_DIR, "07_imputation_mask.csv"))
write_csv(mnar_audit, file.path(cfg$DATA_DIR, "08_mnar_imputation_audit.csv"))

# Plain-text summary
info <- list(
  n_proteins   = nrow(mat), n_samples = ncol(mat), pct_missing = pct_miss,
  n_complete   = n_comp_prots, n_mar_proteins = n_mar_prots,
  n_mnar_proteins = n_mnar_prots, n_mar_values = mar_miss_vals,
  n_mnar_values = mnar_miss_vals,
  pct_mar_values = round(mar_miss_vals / total_miss_vals * 100, 1),
  classification_method = mar_result$method,
  n_unreliable = sum(!miss_class$imputation_reliable),
  best_method  = best,
  best_nrmse   = round(bench_sum$mean_nrmse[1], 4),
  best_pss     = round(bench_sum$mean_pss[1], 4))
writeLines(paste(names(info), info, sep = " = "),
           file.path(cfg$DATA_DIR, "09_imputation_summary.txt"))
```

### Save intermediates for the report script

The companion script `02_imputation_reports.R` generates the missingness and imputation reports. We save all the data it needs into an intermediates file so it can produce plots without re-running the benchmark.

```{r}
#| label: save-intermediates

# Palettes (self-contained, for intermediates)
PAL_GT    <- c(Young_Pre = scales::alpha("#4393C3", 0.5), Young_Post = "#4393C3",
               Old_Pre   = scales::alpha("#D6604D", 0.5), Old_Post   = "#D6604D")
PAL_MAR   <- c(MAR = "#4393C3", MNAR = "#D6604D")
PAL_CLASS <- c(Complete = "#4DAF4A", MAR = "#4393C3", MNAR = "#D6604D")
PAL_MTYPE <- c(MNAR = "#D6604D", MAR = "#4393C3", Hybrid = "#5AAE61")
PAL_BIN   <- c(low = "#D6604D", mid = "#F4A582", high = "#92C5DE")

saveRDS(list(
  mat = mat, mat_imp = mat_imp, was_na = was_na, ann = ann, meta = meta,
  miss_class = miss_class, miss_by_group = miss_by_group,
  prot_pct = prot_pct, pct_miss = pct_miss,
  mnar_genes = mnar_genes, mnar_audit = mnar_audit,
  bench_sum = bench_sum, bench_df = bench_df,
  bin_sum = bin_sum, ext_sum = ext_sum, top5 = top5, best = best,
  METHOD_TYPE = METHOD_TYPE,
  N_ITER = cfg$N_ITER, MASK_FRAC = cfg$MASK_FRAC,
  n_mar_prots = n_mar_prots, n_mnar_prots = n_mnar_prots,
  mar_miss_vals = mar_miss_vals, mnar_miss_vals = mnar_miss_vals,
  total_miss_vals = total_miss_vals,
  PAL_GT = PAL_GT, PAL_MAR = PAL_MAR, PAL_CLASS = PAL_CLASS,
  PAL_MTYPE = PAL_MTYPE, PAL_BIN = PAL_BIN
), file.path(cfg$DATA_DIR, "00_report_intermediates.rds"))

cat(sprintf("Done: %s (NRMSE %.4f) | %d unreliable proteins\n",
            best, bench_sum$mean_nrmse[1], sum(!miss_class$imputation_reliable)))
```

---

## Output Summary

### Data outputs (`c_data/`)

| File | Description |
|------|-------------|
| `01_imputed.csv` | **Canonical output**: complete expression matrix (0% missing) |
| `01_DAList_imputed.rds` | DAList with imputed data + missingness annotations |
| `02_mar_mnar_classification.csv` | Per-protein classification + reliability flag |
| `03_benchmark_summary.csv` | NRMSE + PSS summary for all 12 methods |
| `04_benchmark_raw_iterations.csv` | Per-iteration results (12 methods x 20 iter) |
| `05_benchmark_extended.csv` | Top 5 methods with per-intensity-bin breakdown |
| `06_benchmark_per_intensity.csv` | Per-intensity-tertile NRMSE for top 5 |
| `07_imputation_mask.csv` | Boolean matrix: TRUE = value was originally missing |
| `08_mnar_imputation_audit.csv` | Pre/post means + Cohen's d for MNAR proteins |
| `09_imputation_summary.txt` | Key statistics in plain text |
| `10_imputation_supp.xlsx` | Supplementary workbook (generated by report script) |

### Report outputs (`b_reports/`)

| Report | Contents |
|--------|----------|
| `01_missingness_report.pdf` | Missingness histogram, classification donut, MAR/MNAR scatter, per-sample bars, per-group heatmap |
| `02_imputation_report.pdf` | Benchmark ranking (NRMSE + PSS), observed vs imputed distributions, MNAR audit, extended diagnostics, per-sample LOOCV |

---

## Validation and Limitations

### Post-imputation checks

The imputation report (`02_imputation_report.pdf`) includes several diagnostic plots:

- **Observed vs imputed histogram:** Imputed values (red) should be shifted slightly left relative to observed (blue), reflecting left-censored imputation for MNAR proteins. No dramatic bimodality or implausible values should appear.
- **MNAR mean shift:** Pre vs post mean scatter should show points near or below the 1:1 line.
- **Per-sample LOOCV:** Samples with high missingness should have modestly higher NRMSE, but not dramatically so.

### Known limitations

1. **MNAR imputation is inherently uncertain.** We are estimating values that were never observed. Imputed values reflect model assumptions, not direct measurements.

2. **Classification uncertainty.** The MAR/MNAR boundary is a modeling choice, not a biological truth. Some proteins near the boundary may be misclassified.

3. **Benchmark only validates MAR accuracy.** NRMSE is computed only on artificially masked MAR values (since we know their true values). The benchmark provides only indirect evidence for MNAR imputation quality.

4. **Imputation sensitivity is checked in Phase 3.** The DEP robustness analysis compares limma results on imputed vs non-imputed data, quantifying how much downstream conclusions depend on the imputation.

Now that we have a complete expression matrix with `r nrow(mat_imp)` proteins across `r ncol(mat_imp)` samples (0% missing), we proceed to Phase 3: Differential Expression Analysis.

---

## Downstream Dependencies

The imputed matrix is consumed by:

- **Phase 3 (DEP):** `01_imputed.csv` for imputation sensitivity analysis (not primary DEP)
- **F4 (Clustering):** FCM temporal clustering on imputed data
- **F5 (WGCNA):** Co-expression network analysis on imputed data

The **non-imputed** matrix (`02_normalized.csv`) remains the primary input for differential expression (Phase 3), which handles missing values per-protein via `na.rm` in limma [@Karpievitch2012].

---

## References

::: {#refs}
:::

---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
