---
title: "YvO Proteomics Pipeline — Phase 2: Imputation"
author: "DTL"
date: "2026-02-26"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# 1. Overview

This phase addresses missing values in the normalized proteomics matrix. DIA-MS
data typically has 10–20% missing values arising from two distinct mechanisms:

1. **Missing At Random (MAR):** Missing due to technical variability (peptide
   interference, chromatographic issues). The probability of being missing does
   NOT depend on the protein's true abundance.
2. **Missing Not At Random (MNAR):** Missing because the protein is below the
   detection limit. Low-abundance proteins are systematically more likely to be
   missing.

Different imputation strategies are optimal for each type. This script:

1. Classifies each protein as MAR, MNAR, or Complete
2. Benchmarks 12 imputation methods (4 MNAR, 5 MAR, 3 Hybrid)
3. Selects the best method by NRMSE + Procrustes SS
4. Applies the winner to produce a complete matrix

**Input:** `01_normalization/c_data/02_normalized.csv` (2,124 proteins × 62
samples, 11.67% missing)

**Output:** `02_Imputation/c_data/01_imputed.csv` (2,124 proteins × 62 samples,
0% missing)

**Script:** `02_Imputation/a_script/YvO_imputation_run.R`

---

# 2. Literature & Methods Justification

## 2.1 Why Not Just Remove Incomplete Proteins?

Complete-case analysis (keeping only proteins with zero missing values) would
discard 1,137 of 2,124 proteins (53.5%), including many biologically important
ones. Proteins with partial missingness often have lower abundance and are
precisely those most likely to show differential expression in aging or
exercise. Imputation preserves statistical power while explicitly modeling the
missing data mechanism.

**Reference:** Webb-Robertson et al. 2015 (J Proteome Res 14:920–930):
comprehensive review showing imputation outperforms complete-case analysis for
MS proteomics.

## 2.2 MAR vs MNAR Classification

The script uses **msImpute EBM** (Hediyeh-zadeh et al. 2023, Mol Cell
Proteomics 22:100477) as the primary classifier. This entropy-based method
examines the information content of the missingness pattern: MAR proteins have
missingness unrelated to their abundance, while MNAR proteins show a
systematic relationship between missingness and low intensity.

**Fallback:** If msImpute EBM fails (e.g., due to a missing `group` argument),
the script uses k-means clustering on standardized (intensity, missingness)
features. Proteins in the low-intensity, high-missingness cluster are
classified as MNAR.

## 2.3 Imputation Methods Benchmarked

| Method | Type | Description | Reference |
|--------|------|-------------|-----------|
| MinProb | MNAR | Sample from low-probability Gaussian tail | Lazar et al. 2016 |
| MinDet | MNAR | Deterministic minimum value | Lazar et al. 2016 |
| QRILC | MNAR | Quantile regression left-censored | imputeLCMD |
| zero | MNAR | Replace with zero (baseline) | — |
| knn | MAR | k-nearest neighbors | Troyanskaya et al. 2001 |
| bpca | MAR | Bayesian PCA | Oba et al. 2003 |
| RF | MAR | Random Forest (missForest) | Stekhoven & Bühlmann 2012 |
| SVD | MAR | SVD-based imputation | pcaMethods |
| imputePCA | MAR | Regularized PCA | missMDA |
| mix_bpca_QRILC | Hybrid | bpca (MAR) + QRILC (MNAR) | Borcherding et al. 2021 |
| mix_knn_QRILC | Hybrid | knn (MAR) + QRILC (MNAR) | — |
| mix_RF_QRILC | Hybrid | RF (MAR) + QRILC (MNAR) | — |

**Note:** One MNAR variant per hybrid because the benchmark masks MAR proteins
only — the MNAR component does not affect NRMSE under this design.

**Why hybrid methods?** Lazar et al. 2016 (J Proteome Res 15:1116–1125) and
Borcherding et al. 2021 (Brief Bioinform 22:bbaa112) showed that applying MAR
methods to MAR proteins and MNAR methods to MNAR proteins ("hybrid" or "SFI"
imputation) consistently outperforms using a single method for all proteins.

## 2.4 Benchmark Design

- **20 iterations** with different random masks
- **10% masking rate**: In each iteration, 10% of observed values in MAR
  proteins are artificially set to NA, then imputed
- **Two metrics:**
  - **NRMSE** (Normalized Root Mean Squared Error): Measures point-wise accuracy
    of imputed values compared to the known truth. Lower = better.
  - **Procrustes Sum of Squares (PSS)**: Measures how well the imputed matrix
    preserves the global PCA geometry of the original data. Lower = better
    structure preservation.

**Reference:** Jin et al. 2021 (Sci Rep 11:1556) describes the masking-based
NRMSE benchmarking approach.

## 2.5 Procrustes Metric

Standard NRMSE only measures point-wise accuracy but cannot detect if
imputation distorts the global correlation/covariance structure. The Procrustes
metric (from the `vegan` package) compares PCA projections before and after
imputation by finding the best rotation/scaling and measuring residual
discrepancy. A method with low NRMSE but high PSS would be accurate per-value
but distort sample relationships.

## 2.6 Imputation Reliability Flag

Webel et al. 2024 (J Proteome Res 23:1108–1119) showed that downstream DE
performance degrades substantially when proteins exceed ~50% missingness —
imputed values dominate the estimated means and inflate or deflate fold
changes unpredictably. The script flags each protein with
`imputation_reliable = TRUE` if the protein is Complete or has <50%
missingness. Proteins flagged as unreliable (`FALSE`) are retained in the
matrix but should be interpreted with extra caution in differential expression
and network analyses.

---

# 3. Data Flow

```
02_normalized.csv (2124 × 62, 11.67% missing)
    │
    ├── MAR/MNAR classification (msImpute EBM or k-means fallback)
    │   ├── Complete: 987 proteins (0% missing)
    │   ├── MAR: 705 proteins
    │   └── MNAR: 432 proteins
    │
    ├── 12-method benchmark (20 iterations, NRMSE + PSS)
    │   └── Winner selected
    │
    └── Apply best method → 01_imputed.csv (2124 × 62, 0% missing)
```

---

# 4. Code Walkthrough

## 4.1 MAR/MNAR Classification

```{r}
#| label: mar-mnar-code
#| eval: false

# Primary: msImpute EBM (entropy-based)
mar_result <- tryCatch({
  feat <- msImpute::selectFeatures(mat[has_na,], method = "ebm",
                                   group = meta$Group_Time)
  list(mar_genes = feat, method = "msImpute_ebm")
}, error = function(e) NULL)

# Fallback: k-means on (intensity, missingness)
if (is.null(mar_result)) {
  km <- kmeans(scale(cbind(mc_sub$mean_intensity, mc_sub$pct_miss)),
               centers = 2, nstart = 25)
  mnar_cl <- which.min(tapply(mc_sub$mean_intensity, km$cluster, mean))
  mar_result <- list(mar_genes = mc_sub$gene[km$cluster != mnar_cl])
}

# Reliability flag (Webel et al. 2024): >50% missing → unreliable
miss_class <- miss_class %>%
  mutate(imputation_reliable = classification == "Complete" | pct_miss < 50)
```

**What this does:** Attempts msImpute EBM first. If it fails (the `group`
argument may not be available in all versions), falls back to k-means
clustering on standardized intensity and missingness. The cluster with lower
mean intensity is classified as MNAR.

**How to interpret the classification scatter plot (Report 1, Panel C):**
- X-axis: mean observed intensity (higher = more abundant)
- Y-axis: % missing
- Blue dots (MAR): scattered across intensity range with moderate missingness
- Red dots (MNAR): concentrated at low intensity with high missingness

## 4.2 Benchmark Loop

```{r}
#| label: benchmark-code
#| eval: false

for (iter in seq_len(N_ITER)) {
  # Mask 10% of observed MAR values
  mar_obs_idx <- which(!is.na(mat) & randna[row(mat)])
  mask_idx <- sample(mar_obs_idx, round(length(mar_obs_idx) * 0.10))
  true_v <- mat[mask_idx]
  mm <- mat; mm[mask_idx] <- NA

  for (nm in names(METHODS)) {
    result <- tryCatch(run_impute(METHODS[[nm]], mm, randna), ...)
    nrmse_val <- nrmse(true_v, result$val[mask_idx])
    pss_val   <- pss_metric(mat, result$val)
  }
}
```

**What this does:** For each of 20 iterations: (1) randomly mask 10% of
observed values from MAR proteins, (2) run all 12 methods, (3) compare imputed
values to known truth using NRMSE and PSS. The method with lowest mean NRMSE
wins.

**Key detail:** Only MAR protein values are masked for benchmarking (not MNAR),
because we have no "ground truth" for MNAR values — by definition, they were
never observed.

## 4.3 Applying the Best Method

```{r}
#| label: apply-best
#| eval: false

mat_imp <- run_impute(METHODS[[best]], mat, randna)
stopifnot(sum(is.na(mat_imp)) == 0)
```

**What this does:** Applies the winning method to the full matrix (including
its original missing values). The `randna` vector tells hybrid methods which
proteins are MAR vs MNAR, so they apply the appropriate sub-method.

## 4.4 MNAR Audit

```{r}
#| label: mnar-audit-code
#| eval: false

mnar_audit <- tibble(
  gene = mnar_genes,
  pre_mean  = rowMeans(mat[mnar_genes, ], na.rm = TRUE),
  post_mean = rowMeans(mat_imp[mnar_genes, ]),
  pre_sd    = apply(mat[mnar_genes, ], 1, sd, na.rm = TRUE),
  pct_miss  = prot_pct[mnar_genes],
  shift     = post_mean - pre_mean,
  effect_d  = shift / pre_sd,
  imputation_reliable = pct_miss < 50
)
```

**What this does:** For each MNAR protein, compares the mean intensity before
and after imputation. MNAR methods typically impute values at the lower tail
of the distribution, so `post_mean` should be lower than `pre_mean` (negative
shift). Large positive shifts would indicate problematic imputation.

**What to look for in Report 2, Page 3:**
- **Left panel:** Pre vs post mean scatter. Points below the 1:1 line indicate
  imputed values pulled the mean downward (expected for MNAR methods).
- **Right panel:** Shift vs % missing. Proteins with higher missingness should
  show larger negative shifts (more values imputed from the low tail).

## 4.5 Extended Diagnostics

```{r}
#| label: extended-diagnostics-code
#| eval: false

# Per-intensity-bin NRMSE for top 5 methods
mar_means  <- rowMeans(mat[mar_genes, ], na.rm = TRUE)
bin_breaks <- quantile(mar_means, probs = c(0, 1/3, 2/3, 1))
mar_bins   <- cut(mar_means, breaks = bin_breaks,
                  labels = c("low", "mid", "high"), include.lowest = TRUE)

# For each iteration × top-5 method: compute NRMSE within each intensity bin
# → bin_sum: method × bin → mean_nrmse, sd_nrmse
```

**What this does:** Splits MAR proteins into three intensity tertiles (low,
mid, high abundance) and computes per-bin NRMSE for each of the top 5 methods.
This reveals whether a method struggles at particular intensity ranges — e.g.,
some methods are accurate for mid-abundance proteins but poor at the low end
where values border the MNAR regime.

**What to look for in Report 2, Page 4:**
- **Panel A (grouped bar chart):** Per-bin NRMSE for each method. Ideally,
  the winning method performs well across all three tertiles.
- **Panel B (summary table):** Overall NRMSE, PSS, and per-bin breakdown for
  the top 5 methods.

---

# 5. Output Files

## 5.1 Data Outputs (`c_data/`)

| File | Description |
|------|-------------|
| `01_imputed.csv` | Complete expression matrix (no missing values) |
| `01_DAList_imputed.rds` | R DAList object with imputed data + metadata |
| `02_mar_mnar_classification.csv` | Per-protein MAR/MNAR/Complete classification + `group_miss_pval` + reliability flag |
| `03_benchmark_summary.csv` | NRMSE + PSS summary for all 12 methods |
| `04_benchmark_raw_iterations.csv` | Raw per-iteration NRMSE + PSS for all 12 methods × 20 iterations |
| `05_benchmark_extended.csv` | Extended diagnostics: NRMSE + PSS + per-intensity-bin NRMSE for top 5 |
| `06_benchmark_per_intensity.csv` | Per-intensity-tertile NRMSE for top 5 methods |
| `07_imputation_mask.csv` | Boolean matrix indicating which values were imputed (TRUE = was NA) |
| `08_mnar_imputation_audit.csv` | Per-MNAR-protein pre/post imputation means + Cohen's d |
| `09_imputation_summary.txt` | Key statistics (best method, NRMSE, counts) |
| `imputation_supp.xlsx` | Supplementary Excel workbook (7 sheets, mirrors CSV outputs) |

## 5.2 Report Outputs (`b_reports/`)

| Report | Contents |
|--------|----------|
| `01_missingness_report.pdf` | 6-panel missingness characterization: histogram, classification donut, MAR/MNAR scatter, per-sample bars, intensity distributions, group-level heatmap |
| `02_imputation_report.pdf` | 4-page report: benchmark results (NRMSE + PSS), observed vs imputed distributions, MNAR audit, extended diagnostics |

---

# 6. Results & Interpretation

## 6.1 Benchmark Results

```{r}
#| label: benchmark-results

bench <- read.csv("../c_data/03_benchmark_summary.csv")
knitr::kable(bench[1:5, ], caption = "Top 5 methods by NRMSE", digits = 4)
```

**Interpretation:** Hybrid methods (mixing a MAR method with an MNAR method)
often outperform single-method approaches (Lazar et al. 2016; Borcherding et
al. 2021), but at low overall missingness (~12%), pure MAR methods may match
or beat hybrids because even MNAR-classified proteins retain enough observed
values for structure-preserving algorithms to fit reliable models. The best
method by NRMSE is applied to produce the final imputed matrix.

## 6.2 MAR/MNAR Classification

```{r}
#| label: classification-results

class_df <- read.csv("../c_data/02_mar_mnar_classification.csv")
cat(sprintf("Classification summary:\n"))
print(table(class_df$classification))
```

**Expected pattern:**
- ~40–50% Complete (zero missing)
- ~25–35% MAR (moderate missingness, intensity-independent)
- ~15–25% MNAR (low abundance, high missingness)

## 6.3 MNAR Imputation Audit

```{r}
#| label: mnar-audit-results

audit <- read.csv("../c_data/08_mnar_imputation_audit.csv")
cat(sprintf("MNAR audit: %d proteins\n", nrow(audit)))
cat(sprintf("Median shift: %.3f log2 units\n", median(audit$shift)))
cat(sprintf("Range: [%.3f, %.3f]\n", min(audit$shift), max(audit$shift)))
cat(sprintf("Reliable (< 50%% missing): %d (%.1f%%)\n",
            sum(audit$imputation_reliable),
            100 * mean(audit$imputation_reliable)))
```

**Expected:** Negative median shift (MNAR imputed values are lower than the
observed mean, consistent with left-censored imputation). Proteins with
>50% missing should be treated with caution in downstream analyses.

---

# 7. Validation & Sanity Checks

## 7.1 Post-Imputation Distributions

The observed vs imputed histogram (Report 2, Page 2) should show:
- Imputed values (red) shifted slightly left relative to observed (blue)
- No dramatic bimodality or implausible values
- Per-group density overlays should maintain group-specific distribution shapes

## 7.2 Known Limitations

1. **MNAR imputation is inherently uncertain:** We are estimating values that
   were never observed. The imputed values reflect our model assumptions (e.g.,
   Gaussian left tail), not direct measurements. Downstream analyses should be
   interpreted with this caveat, especially for MNAR-classified proteins.

2. **Classification uncertainty:** The MAR/MNAR boundary is a modeling choice,
   not a biological truth. Some proteins near the boundary may be misclassified.
   The k-means fallback (when msImpute fails) uses a hard boundary that may
   not capture the gradual transition between mechanisms.

3. **Benchmark limitation:** NRMSE is computed only on artificially masked MAR
   values, not on true MNAR values (which are unknown). The benchmark validates
   accuracy for MAR proteins but provides only indirect evidence for MNAR
   imputation quality.

4. **Sensitivity to classification:** The pipeline includes an imputation
   sensitivity analysis in Phase 3 (DEP) that compares limma results on
   imputed vs non-imputed data, quantifying how much downstream conclusions
   depend on the imputation.

5. **Group-stratified missingness:** The script now includes a Fisher exact
   test for each MNAR protein checking whether missingness is uniformly
   distributed across the 4 experimental groups (Young_Pre, Young_Post,
   Old_Pre, Old_Post). Proteins with significant group-specific missingness
   (p < 0.05) are flagged via the `group_miss_pval` column in
   `02_mar_mnar_classification.csv`. These may warrant extra caution, as their
   missingness could be confounded with biological signal.

6. **PSS metric fix (2026-02-26):** The Procrustes Sum of Squares metric now
   computes PCA on complete-case rows only (samples with no missing values in
   the original matrix), avoiding NA propagation in `prcomp`. This produces
   valid PSS values for all benchmarked methods.

7. **bpca-as-winner disclosure:** When the benchmark winner is a pure MAR
   method (e.g., bpca), the MAR/MNAR classification does not affect imputed
   values — all proteins receive the same imputation. The classification is
   retained for reporting and the MNAR audit. At low overall missingness
   (~12%), MNAR-classified proteins retain enough observed values for EM-based
   methods to fit reliable models, which may explain why structure-preserving
   methods outperform left-censored approaches (Hediyeh-zadeh et al. 2023,
   MCP 22:100477).

8. **Imputation reliability flag:** Proteins with >50% missingness are flagged
   as `imputation_reliable = FALSE` in both `02_mar_mnar_classification.csv` and
   the DAList annotation. This threshold follows Webel et al. 2024 (J Proteome
   Res 23:1108–1119), who showed that DE performance degrades when imputed
   values dominate the intensity estimates. The flag is propagated to Phase 3
   (DEP) and downstream figures but does not remove proteins from the matrix.

---

# 8. Downstream Dependencies

The imputed matrix is used by:
- **Phase 3 (DEP):** `01_imputed.csv` for imputation sensitivity analysis
- **Phase 3 (DEP):** `01_DAList_imputed.rds` for imputed limma comparison
- **F3 (Reversal tests):** Melov permutation uses imputed group means
- **F4 (Clustering):** FCM temporal clustering on imputed data
- **F5 (WGCNA):** Co-expression network on imputed data

The **non-imputed** matrix (`02_normalized.csv`) is used for the primary DEP
analysis (Phase 3), which handles missing values via `na.rm` in limma.

---

# 9. Session Information

```{r}
#| label: session-info
sessionInfo()
```
