---
title: "YvO Proteomics Pipeline — Phase 3: Differential Expression"
author: "DTL"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
bibliography: ../../references.bib
---

# Overview

In Phases 1 and 2, we produced a normalized expression matrix (2,124 proteins x 62 samples) and a companion imputed matrix. Now we answer the central biological question: **which proteins change significantly between conditions?**

This is **differential expression analysis** — the statistical comparison of protein abundance across groups. We use **limma** [@Ritchie2015; @Smyth2004], a mature and widely-used framework that applies empirical Bayes moderation to stabilize variance estimates, especially important when sample sizes are modest (n = 15--16 per group).

Our study design defines **five contrasts** (comparisons):

| Contrast | Definition | Biological question |
|----------|-----------|-------------------|
| Training_Young | Young_Post - Young_Pre | How does training change the young proteome? |
| Training_Old | Old_Post - Old_Pre | How does training change the old proteome? |
| Aging | Old_Pre - Young_Pre | What differs between young and old at baseline? |
| Interaction | (Old_Post - Old_Pre) - (Young_Post - Young_Pre) | Does the training response differ by age? |
| Reversal | (Old_Post - Old_Pre) - (Old_Pre - Young_Pre) | Does training in older adults reverse age-related changes? |

A critical design choice: we run limma on the **non-imputed** normalized matrix [@Karpievitch2012]. Limma handles missing values per-protein (dropping NAs from each model fit), which avoids introducing imputation artifacts into our statistical tests. The imputed matrix is used only for visualization and as a sensitivity check.

> **Note:** This tutorial covers three scripts: `01_run_dep.R` (model fitting and results extraction), `02_dep_reports.R` (visualization), and `03_dep_robustness.R` (blunting diagnostics, bootstrap CIs, power analysis, imputation sensitivity). All computations, thresholds, and outputs are preserved exactly as they appear in those scripts.

**Input:** `01_normalization/c_data/02_normalized.csv` + `03_DAList_normalized.rds`

**Output:** `03_DEP/c_data/03_combined_results.csv` (wide-format, all contrasts), per-contrast CSVs, summary Excel

---

## Prerequisites

### Required packages

```{r}
#| label: setup

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(purrr)
library(proteoDA)
library(openxlsx)
```

Key packages:

- **proteoDA** [@Thurman2023] — Provides `DAList` structure, `add_design()`, `add_contrasts()`, `fit_limma_model()`, and `extract_DA_results()`. These functions wrap the limma workflow into a consistent API that keeps the data, design, and results together.
- **limma** [@Ritchie2015] — The statistical engine underneath. It uses empirical Bayes methods to borrow information across proteins, shrinking per-protein variance estimates toward a common prior. This dramatically improves power when sample sizes are small — exactly our situation (15--16 subjects per age group).
- **openxlsx** — For writing multi-sheet Excel workbooks (supplementary tables).

### Configuration

```{r}
#| label: config

set.seed(42)

cfg <- list(
  norm_csv    = "01_normalization/c_data/02_normalized.csv",
  norm_rds    = "01_normalization/c_data/03_DAList_normalized.rds",
  data_dir    = "03_DEP/c_data",
  report_dir  = "03_DEP/b_reports",
  per_dir     = "03_DEP/c_data/04_per_contrast_results",
  proteoDA_dir = "03_DEP/b_reports/01_proteoDA",
  pval_thresh = 0.10,     # FDR threshold
  lfc_thresh  = 0,        # no log-fold-change cutoff
  adj_method  = "BH",     # Benjamini-Hochberg
  pi_thresh   = 0.05      # Pi-score threshold
)

dir.create(cfg$data_dir,     recursive = TRUE, showWarnings = FALSE)
dir.create(cfg$per_dir,      recursive = TRUE, showWarnings = FALSE)
dir.create(cfg$proteoDA_dir, recursive = TRUE, showWarnings = FALSE)
```

**Why FDR = 0.10?** This is an **exploratory** proteomics study with modest sample sizes (n = 15--16). @BenjaminiHochberg1995 used 0.10 in their original examples, and @Choi2008 recommend FDR of 0--15% for label-free proteomics. We also use the **Pi-score** [@Xiao2014] as a secondary, effect-size-weighted significance criterion (see below).

---

## Step 1: Load Data and Build Metadata

We load the normalized (non-imputed) matrix and extract metadata from the DAList object.

```{r}
#| label: load-data

df <- read_csv(cfg$norm_csv, show_col_types = FALSE)

ann_cols   <- c("uniprot_id", "protein", "gene", "description")
ann        <- df[, ann_cols]
samp_names <- setdiff(names(df), ann_cols)
mat        <- as.matrix(df[, samp_names])
rownames(mat) <- ann$uniprot_id

cat(sprintf("Loaded: %d proteins x %d samples | missing: %d (%.1f%%)\n",
            nrow(mat), ncol(mat), sum(is.na(mat)),
            100 * sum(is.na(mat)) / length(mat)))

# Canonical metadata from normalisation DAList (not regex-derived)
dal_norm <- readRDS(cfg$norm_rds)
dal_meta <- as.data.frame(dal_norm$metadata)
meta <- tibble(
  sample_id = dal_meta$Col_ID,
  age       = dal_meta$Group,
  time      = dal_meta$Timepoint,
  group     = dal_meta$Group_Time,
  # Col_ID prefix as subject key (Subject_ID is NOT globally unique —
  # 8 IDs are shared between Young/Old cohorts, different individuals)
  subject   = sub("_(Pre|Post)$", "", dal_meta$Col_ID)
)
meta$age   <- factor(meta$age,  levels = c("Young", "Old"))
meta$time  <- factor(meta$time, levels = c("Pre", "Post"))
meta$group <- factor(meta$group,
                     levels = c("Young_Pre", "Young_Post", "Old_Pre", "Old_Post"))

stopifnot(setequal(colnames(mat), meta$sample_id))
```

> **Note:** We derive the subject key from `Col_ID` using `sub("_(Pre|Post)$", "", Col_ID)`, not from the `Subject_ID` column. This is because `Subject_ID` is **not globally unique** — 8 IDs are shared between the Young and Old cohorts (but they represent different individuals). The Col_ID-based key is unique.

---

## Step 2: Create DAList

We package the intensity matrix, annotations, and metadata into a proteoDA `DAList` object.

```{r}
#| label: create-dalist

meta_df <- as.data.frame(meta)
rownames(meta_df) <- meta$sample_id

dal <- DAList(
  data       = mat,
  annotation = as.data.frame(ann),
  metadata   = meta_df,
  tags       = list(norm_method = "cycloess")
)
```

---

## Step 3: Define the Statistical Design

Our study has a **repeated-measures** structure: each subject contributes both a Pre and a Post sample. We need to account for the **within-subject correlation** — the fact that two samples from the same person are more similar to each other than to samples from different people.

The design formula `~ 0 + group + (1 | subject)` specifies:

- `0 + group` — a cell-means model with one coefficient per group (Young_Pre, Young_Post, Old_Pre, Old_Post). The `0` removes the intercept so each coefficient directly estimates the mean for its group.
- `(1 | subject)` — a random intercept for subject, which tells proteoDA to use `limma::duplicateCorrelation()` to estimate and account for the within-subject correlation [@Smyth2005].

```{r}
#| label: design

dal <- add_design(dal, "~ 0 + group + (1 | subject)")
colnames(dal$design$design_matrix) <- gsub("^group", "",
                                            colnames(dal$design$design_matrix))
```

The `duplicateCorrelation()` function [@Smyth2005, Section 9.7] estimates a single consensus correlation across all proteins. This value (typically 0.3--0.6 for proteomics) captures how much of each protein's variance is explained by between-subject differences versus within-subject (Pre/Post) differences. Accounting for this correlation improves power for within-subject contrasts (Training_Young, Training_Old) and produces correct standard errors.

---

## Step 4: Define Contrasts

**Contrasts** are the specific comparisons we want to test. Each contrast is a linear combination of the group means.

```{r}
#| label: contrasts

dal <- add_contrasts(dal, contrasts_vector = c(
  "Training_Young = Young_Post - Young_Pre",
  "Training_Old = Old_Post - Old_Pre",
  "Aging = Old_Pre - Young_Pre",
  "Interaction = (Old_Post - Old_Pre) - (Young_Post - Young_Pre)",
  "Reversal = (Old_Post - Old_Pre) - (Old_Pre - Young_Pre)"
))
```

A few notes on these contrasts:

- **Training_Young** and **Training_Old** are simple paired differences (Post minus Pre within each age group). These directly test whether resistance training changes the proteome.
- **Aging** compares baseline (Pre) samples between age groups. This captures the proteomic signature of aging.
- **Interaction** tests whether the training effect *differs* between age groups — formally, it is a difference-of-differences.
- **Reversal** tests whether training in older adults moves their proteome *toward* the young baseline. It is defined as `(Old_Post - Old_Pre) - (Old_Pre - Young_Pre)` — the training effect in old minus the aging effect. This is a non-standard but biologically motivated contrast inspired by @Melov2007. It is **not orthogonal** to Training_Old and Aging (they share terms), which should be acknowledged in the methods.

> **Note:** The Interaction contrast typically yields very few (0--1) FDR-significant proteins at our sample size (n ~ 15/group). This is expected — detecting interaction effects requires substantially more power than detecting main effects. Similarly, Training_Old yields 0 FDR hits, consistent with an attenuated training response in older adults.

---

## Step 5: Fit the Model and Extract Results

This is the core statistical step. `fit_limma_model()` internally calls `limma::lmFit()` (linear model fitting) and `limma::eBayes()` (empirical Bayes moderation). The moderation step is what makes limma powerful for small samples: it borrows information across all 2,124 proteins to produce more stable variance estimates for each individual protein.

```{r}
#| label: fit-model

dal <- fit_limma_model(dal)

# Report within-subject correlation
within_cor <- dal$eBayes_fit$correlation %||%
  dal$tags$duplicate_correlation %||% NA_real_
if (!is.na(within_cor)) cat(sprintf("Within-subject correlation: %.3f\n", within_cor))
```

Now we extract results using the configured FDR threshold:

```{r}
#| label: extract-results

dal <- extract_DA_results(dal,
                          pval_thresh = cfg$pval_thresh,
                          lfc_thresh  = cfg$lfc_thresh,
                          adj_method  = cfg$adj_method)
```

The `extract_DA_results()` function runs `limma::topTable()` for each contrast, applying Benjamini-Hochberg correction [@BenjaminiHochberg1995] for multiple testing. The `adj.P.Val` column in the results is the FDR-adjusted p-value.

---

## Step 6: Save the Fitted DAList

```{r}
#| label: save-dalist

saveRDS(dal, file.path(cfg$data_dir, "01_limma_DAList.rds"))
```

This RDS file contains everything: the data, metadata, design, contrasts, and the fitted limma model. It is consumed by the report script and by figure scripts downstream.

---

## Step 7: Generate proteoDA Reports

proteoDA can automatically generate standard limma report PDFs (volcano plots, MD plots, p-value histograms) for each contrast:

```{r}
#| label: proteoDA-reports

tryCatch(
  write_limma_reports(dal, output_dir = cfg$proteoDA_dir, overwrite = TRUE),
  error = function(e) cat(sprintf("write_limma_reports: %s\n", conditionMessage(e)))
)

tryCatch(
  write_limma_plots(dal,
                    grouping_column = "group",
                    output_dir      = cfg$proteoDA_dir,
                    table_columns   = c("uniprot_id", "gene", "protein"),
                    title_column    = "gene",
                    overwrite       = TRUE),
  error = function(e) cat(sprintf("write_limma_plots: %s\n", conditionMessage(e)))
)
```

These are standard diagnostic plots. The p-value histograms are particularly informative: a uniform distribution suggests no signal (no true positives), while a spike near zero indicates genuine differential expression.

---

## Step 8: Build Results and Compute Pi-scores

### What is the Pi-score?

The **Pi-score** [@Xiao2014] integrates statistical significance with effect size into a single metric:

$$\text{Pi} = p^{|\log_2 \text{FC}|}$$

where $p$ is the raw p-value and $|\log_2 \text{FC}|$ is the absolute log-fold change. A protein needs both a small p-value *and* a large fold change to achieve a low Pi-score. Our threshold of Pi < 0.05 (equivalent to the original pi > 1.3 in Xiao et al.'s formulation) provides an effect-size-weighted alternative to pure FDR filtering.

**Why use Pi-scores alongside FDR?** FDR controls the false discovery rate but can flag proteins with statistically significant but biologically tiny fold changes (especially with large sample sizes). The Pi-score naturally penalizes small effects. Conversely, proteins with large fold changes but moderate p-values (just missing the FDR threshold) may still be biologically important — Pi can capture these.

```{r}
#| label: build-results

contrast_names <- names(dal$results)
ann_df <- as.data.frame(dal$annotation)

# Build results with Pi-scores for each contrast
results_list <- lapply(contrast_names, function(cname) {
  dal$results[[cname]] |>
    rownames_to_column("uniprot_id") |>
    left_join(ann_df, by = "uniprot_id") |>
    mutate(
      pi_score = P.Value ^ abs(logFC),          # Pi = p^|logFC|
      sig_pi = case_when(
        pi_score < cfg$pi_thresh & logFC > 0 ~  1L,   # significant up
        pi_score < cfg$pi_thresh & logFC < 0 ~ -1L,   # significant down
        TRUE ~ 0L)                                      # not significant
    ) |>
    select(-any_of(c("sig.PVal", "sig.FDR"))) |>
    mutate(contrast = cname)
})
names(results_list) <- contrast_names
```

---

## Step 9: Write Per-Contrast CSVs

Each contrast gets its own CSV with full protein-level results (annotation + expression values + statistics):

```{r}
#| label: per-contrast-csv

data_df <- as.data.frame(dal$data)

for (cname in contrast_names) {
  res <- results_list[[cname]]
  out <- ann_df |>
    bind_cols(data_df) |>
    left_join(
      res |> select(uniprot_id, logFC, CI.L, CI.R, average_intensity,
                     t, B, P.Value, adj.P.Val, pi_score, sig_pi),
      by = "uniprot_id"
    )
  write_csv(out, file.path(cfg$per_dir, paste0(cname, ".csv")))
}
```

---

## Step 10: Build Combined Results (Wide Format)

The combined results file merges all contrasts into a single wide-format CSV. Each statistic (logFC, P.Value, adj.P.Val, pi_score, etc.) gets a contrast suffix (e.g., `logFC_Aging`, `pi_score_Training_Young`).

```{r}
#| label: combined-results

base_df <- bind_cols(
  ann_df |> select(any_of(c("uniprot_id", "protein", "gene", "description"))),
  data_df
)

# Add per-contrast stats as suffixed columns
for (cname in contrast_names) {
  res <- results_list[[cname]]
  stat_cols <- res |>
    select(uniprot_id, logFC, CI.L, CI.R, average_intensity, t, B,
           P.Value, adj.P.Val, pi_score, sig_pi)
  names(stat_cols)[-1] <- paste0(names(stat_cols)[-1], "_", cname)
  base_df <- left_join(base_df, stat_cols, by = "uniprot_id")
}

write_csv(base_df, file.path(cfg$data_dir, "03_combined_results.csv"))
```

> **Note:** This is a **wide-format** file. Column names are like `logFC_Aging`, `adj.P.Val_Training_Young`, etc. Downstream figure scripts expect this format — do not reshape it.

---

## Step 11: Build DA Summary

```{r}
#| label: da-summary

da_summary <- map_dfr(contrast_names, function(cname) {
  res <- results_list[[cname]]
  bind_rows(
    tibble(contrast = cname, type = "up",
           sig.PVal   = sum(res$P.Value < cfg$pval_thresh & res$logFC > 0, na.rm = TRUE),
           sig.FDR    = sum(res$adj.P.Val < cfg$pval_thresh & res$logFC > 0, na.rm = TRUE),
           pval_thresh = cfg$pval_thresh, lfc_thresh = cfg$lfc_thresh,
           p_adj_method = cfg$adj_method,
           sig.Pi     = sum(res$sig_pi == 1, na.rm = TRUE),
           sig.FDR.05 = sum(res$adj.P.Val < 0.05 & res$logFC > 0, na.rm = TRUE),
           sig.FDR.10 = sum(res$adj.P.Val < 0.10 & res$logFC > 0, na.rm = TRUE)),
    tibble(contrast = cname, type = "down",
           sig.PVal   = sum(res$P.Value < cfg$pval_thresh & res$logFC < 0, na.rm = TRUE),
           sig.FDR    = sum(res$adj.P.Val < cfg$pval_thresh & res$logFC < 0, na.rm = TRUE),
           pval_thresh = cfg$pval_thresh, lfc_thresh = cfg$lfc_thresh,
           p_adj_method = cfg$adj_method,
           sig.Pi     = sum(res$sig_pi == -1, na.rm = TRUE),
           sig.FDR.05 = sum(res$adj.P.Val < 0.05 & res$logFC < 0, na.rm = TRUE),
           sig.FDR.10 = sum(res$adj.P.Val < 0.10 & res$logFC < 0, na.rm = TRUE)),
    tibble(contrast = cname, type = "nonsig",
           sig.PVal   = sum(res$P.Value >= cfg$pval_thresh, na.rm = TRUE),
           sig.FDR    = sum(res$adj.P.Val >= cfg$pval_thresh, na.rm = TRUE),
           pval_thresh = cfg$pval_thresh, lfc_thresh = cfg$lfc_thresh,
           p_adj_method = cfg$adj_method,
           sig.Pi     = sum(res$sig_pi == 0, na.rm = TRUE),
           sig.FDR.05 = sum(!res$adj.P.Val < 0.05, na.rm = TRUE),
           sig.FDR.10 = sum(!res$adj.P.Val < 0.10, na.rm = TRUE))
  )
})

write_csv(da_summary, file.path(cfg$data_dir, "02_DA_summary.csv"))
```

---

## Step 12: Write Results Excel Workbook

```{r}
#| label: results-excel

wb_results <- createWorkbook()
for (cname in contrast_names) {
  addWorksheet(wb_results, cname)
  writeData(wb_results, cname, results_list[[cname]])
}
addWorksheet(wb_results, "DA_Summary")
writeData(wb_results, "DA_Summary", da_summary)
saveWorkbook(wb_results, file.path(cfg$data_dir, "05_results.xlsx"), overwrite = TRUE)

cat(sprintf("Done: 01_run_dep.R — %d contrasts -> %s/\n",
            length(contrast_names), cfg$data_dir))
```

---

## Results and Interpretation

### DEP counts

```{r}
#| label: dep-counts

# Summarize DEP counts per contrast
dep_counts <- map_dfr(contrast_names, function(cname) {
  res <- results_list[[cname]]
  tibble(
    Contrast  = cname,
    `FDR < 0.10` = sum(res$adj.P.Val < 0.10, na.rm = TRUE),
    `Pi < 0.05`  = sum(abs(res$sig_pi) == 1, na.rm = TRUE)
  )
})
knitr::kable(dep_counts, caption = "Differentially expressed proteins by contrast")
```

Key observations:

- **Aging** shows the largest signal — hundreds of proteins differ between Young and Old at baseline.
- **Training_Young** has a substantial training response.
- **Training_Old** has far fewer DEPs — a 5--6x blunting compared to Training_Young, consistent with attenuated training response in older adults.
- **Interaction** yields very few FDR-significant proteins, reflecting the low power for detecting age x training interactions at this sample size.
- **Reversal** captures proteins where training in old subjects moves their expression back toward young baseline levels.

### Within-subject correlation

The `duplicateCorrelation()` estimate (reported above) captures how correlated Pre and Post measurements are within the same subject. Values of 0.3--0.6 are typical for proteomics. This correlation is factored into all standard error calculations, improving power for within-subject contrasts (Training_Young, Training_Old).

---

## Robustness Analyses (03_dep_robustness.R)

The robustness script performs four additional analyses to strengthen confidence in the results. We summarize them here conceptually; the full code is in `03_dep_robustness.R`.

### Blunting diagnostics

We formally test whether the training response is attenuated in older adults by comparing `|logFC|` distributions between Training_Young and Training_Old:

- **Kolmogorov-Smirnov test** — distributional shape/location difference
- **Fligner-Killeen test** — variance heterogeneity (robust to non-normality) [@Conover1999]
- **Wilcoxon signed-rank test** — paired location shift (same proteins, different contrasts)
- **Cliff's delta** — non-parametric effect size (positive = young responds more)

### Bootstrap confidence intervals

Median |logFC| with 95% BCa bootstrap CIs [@Efron1993] per contrast, providing a global measure of effect magnitude.

### Power analysis

Minimum detectable logFC at 80% power (alpha = 0.10) using `pwr.t.test()` [@Cohen1988]. These are conservative lower bounds — limma's moderated t has higher power than a standard t-test because of empirical Bayes variance shrinkage.

### Imputation sensitivity

Spearman correlation of t-statistics between non-imputed (primary) and imputed limma fits. High correlation (rho > 0.93) across all contrasts demonstrates that our conclusions are robust to whether we impute or not — validating the non-imputed approach [@Karpievitch2012].

---

## Output Summary

### Data outputs (`c_data/`)

| File | Description |
|------|-------------|
| `01_limma_DAList.rds` | **Canonical**: fitted DAList with limma model |
| `02_DA_summary.csv` | DEP counts by contrast, direction, and criterion |
| `03_combined_results.csv` | **Canonical**: wide-format results (all contrasts) |
| `04_per_contrast_results/` | Per-contrast CSVs with full stats |
| `05_results.xlsx` | Results workbook (one sheet per contrast + summary) |
| `06_blunting_diagnostics.csv` | Blunting test statistics |
| `07_effect_size_bootstrap.csv` | Bootstrap CIs for median |logFC| |
| `08_power_analysis.csv` | Minimum detectable effect sizes |
| `09_imputation_sensitivity.csv` | Non-imputed vs imputed correlation |
| `10_DEP_supplementary.xlsx` | Supplementary workbook (11 sheets) |

### Report outputs (`b_reports/`)

| Report | Description |
|--------|-------------|
| `01_proteoDA/` | Standard limma diagnostic plots (volcano, MD, p-value histograms) |
| `02_dep_overview.pdf` | DEP summary bar chart + outlier sensitivity table |
| `03_contrast_summaries/` | Per-contrast 3-panel volcano + top-25 table |

---

## Validation and Limitations

### Expected behavior

- **P-value histograms** should show a spike near zero for contrasts with real signal (Aging, Training_Young) and a more uniform shape for low-signal contrasts (Interaction, Training_Old).
- **Volcano plots** should show symmetric clouds with labeled extremes.
- **Outlier sensitivity** (in `02_dep_overview.pdf` page 2) should show high Pearson/Spearman correlation between full (64-sample) and reduced (62-sample) analyses, confirming that outlier removal did not distort the results.

### Known limitations

1. **FDR 0.10 is permissive.** We use it because this is an exploratory study with modest sample sizes. The Pi-score provides a secondary filter that requires both significance and meaningful effect size.

2. **The Reversal contrast is non-orthogonal** to Training_Old and Aging. They share terms (`Old_Pre` appears in all three). This means the Reversal contrast is not statistically independent — interpret with caution and pair with visualizations (RRHO2, concordance scatter) in the figure scripts.

3. **Low power for Interaction.** Detecting interaction effects (age x training) requires ~4x the sample size needed for main effects. Our 0--1 FDR-significant interaction proteins reflect power limitations, not necessarily absence of biological interaction.

4. **Non-imputed analysis.** Running limma on non-imputed data means each protein's model is fit on a slightly different subset of samples (those with observed values). This is statistically valid but means protein-to-protein comparisons should consider the effective sample size.

The results from this phase feed into all downstream figures (F0--F6) and form the statistical foundation of the manuscript.

---

## Downstream Dependencies

The combined results file (`03_combined_results.csv`) and fitted DAList (`01_limma_DAList.rds`) are consumed by:

- **F1:** DEP count bar plots, upset plots, GSEA
- **F2:** Training-young and training-old volcano plots, concordance scatter, RRHO2
- **F3:** Aging and reversal volcano plots, reversal analysis
- **F4:** FCM clustering (uses imputed data, but references DEP classification)
- **F5:** WGCNA module-trait correlations (uses imputed data + DEP context)
- **F6:** Age discrimination, plasticity, hub-gene analyses

---

## References

::: {#refs}
:::

---

## Session Information

```{r}
#| label: session-info

sessionInfo()
```
