---
title: "YvO Proteomics Pipeline — Phase 0: Input Data Inventory"
author: "DTL"
date: "2026-02-25"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# 1. Overview

This document provides a comprehensive inventory and validation of the input data for the **YvO proteomics analysis pipeline** — a study examining the effects of **resistance exercise training** on the **skeletal muscle proteome** of **young** and **older** adults.

## 1.1 Study Design

This is a **2 × 2 repeated-measures factorial design**:

| Factor                     | Levels                             |
|----------------------------|------------------------------------|
| **Age** (between-subjects) | Young (19–26 yr) vs Old (42–69 yr) |
| **Time** (within-subjects) | Pre-training vs Post-training      |

Each participant provided vastus lateralis (VL) muscle biopsies at two timepoints (before and after a resistance training intervention), yielding a **paired** design. This enables estimation of individual-level training responses and statistical tests for: - **Training effects within each age group** (Training_Young, Training_Old) - **Age differences at baseline** (Aging) - **Age × Time interaction** (whether training effects differ by age)

### 1.2 Scientific Context

Resistance exercise training (RET) is a potent stimulus for skeletal muscle protein synthesis, fiber hypertrophy, and strength gains. Whether RET can "reverse" age-related proteomic changes — i.e., shift the older adult proteome toward a younger profile — is a central question in exercise geroscience.

-   Melov et al. 2007 (PMID: 17520024) — resistance exercise partially reverses aging-associated gene expression in human skeletal muscle
-   Hostrup et al. 2022 (eLife 11:e69802) — HIT remodels skeletal muscle proteome; \~3,168 proteins quantified by DIA-MS
-   O'Leary et al. 2025 (Exp Physiol 110:438–453) — skeletal muscle proteomics in young/older women after 8-week RET
-   Robinson et al. 2022 (GeroScience) — study cohort context

### 1.3 Proteomics Platform

Samples were analyzed using **data-independent acquisition mass spectrometry (DIA-MS)**, which provides: - Deep proteome coverage (\~2,800+ proteins) - Improved quantitative reproducibility relative to DDA - Fewer missing values than DDA shotgun approaches

**Reference:** Bruderer et al. 2015 (Mol Cell Proteomics 14:1400–1410)

------------------------------------------------------------------------

# 2. Input Files

The `00_input/` directory contains four files:

```{r}
#| label: file-listing
library(readxl)
library(dplyr)
library(tidyr)
library(knitr)

files <- data.frame(
  File = c("YvO_raw.xlsx", "YvO_meta.xlsx",
           "HPA_skeletal_muscle_annotations.tsv"),
  Size = c("1.2 MB", "15 KB", "3.3 MB"),
  Description = c(
    "Raw protein-level intensities (log2) from DIA-MS",
    "Sample-level metadata (design, demographics, phenotypes)",
    "Human Protein Atlas skeletal muscle annotations for tissue filtering"
  )
)
kable(files, caption = "Input file inventory")
```

------------------------------------------------------------------------

# 3. Raw Proteomics Data (`YvO_raw.xlsx`)

## 3.1 Structure

```{r}
#| label: raw-data-structure
raw <- read_excel("YvO_raw.xlsx")

cat(sprintf("Dimensions: %d proteins × %d columns\n", nrow(raw), ncol(raw)))
cat(sprintf("Annotation columns: %d (uniprot_id, protein, gene, description, n_seq)\n", 5))
cat(sprintf("Sample columns: %d\n", ncol(raw) - 5))
```

| Column | Description |
|----------------------------|--------------------------------------------|
| `uniprot_id` | UniProt accession (e.g., P02768) — protein-level identifier |
| `protein` | Protein name from MaxQuant/DIA-NN output |
| `gene` | HGNC gene symbol |
| `description` | Full protein description |
| `n_seq` | Number of unique peptide sequences supporting identification |
| Sample columns | Raw spectral intesities, one per sample |

## 3.2 Sample Column Naming Convention

Sample columns follow the pattern: `{prefix}_{SubjectID}_{Timepoint}`

```{r}
#| label: sample-columns
sample_cols <- colnames(raw)[-(1:5)]
prefixes <- sub("_S.*", "", sample_cols)
prefix_df <- data.frame(
  Prefix = c("O", "OP", "Y", "YP"),
  Group = c("Old (no supplement)", "Old + supplement", "Young (no supplement)", "Young + supplement"),
  Count = c(sum(prefixes == "O"), sum(prefixes == "OP"),
            sum(prefixes == "Y"), sum(prefixes == "YP"))
)
kable(prefix_df, caption = "Column prefix encoding")
```

**Important:** The `Subject_ID` component (e.g., S05, S06) is **not globally unique**. The same Subject_ID may refer to different individuals in the Young vs Old cohorts, or in the supplement vs non-supplement arms. The full Col_ID (prefix + Subject_ID + timepoint) is the unique sample identifier, and the prefix + Subject_ID combination is the unique **subject** identifier for repeated-measures pairing.

## 3.3 Data Quality Summary

```{r}
#| label: raw-data-quality
mat <- as.matrix(raw[, -(1:5)])
rownames(mat) <- raw$uniprot_id

cat(sprintf("Total cells: %s\n", format(length(mat), big.mark = ",")))
cat(sprintf("Missing values: %d (%.1f%%)\n", sum(is.na(mat)),
            100 * sum(is.na(mat)) / length(mat)))
cat(sprintf("Proteins with zero missing: %d (%.1f%%)\n",
            sum(rowSums(is.na(mat)) == 0),
            100 * sum(rowSums(is.na(mat)) == 0) / nrow(mat)))

# Intensity range
vals <- mat[!is.na(mat)]
cat(sprintf("Intensity range: %.2f – %.2f (raw)\n", min(vals), max(vals)))
cat(sprintf("Median intensity: %.2f\n", median(vals)))
```

------------------------------------------------------------------------

# 4. Sample Metadata (`YvO_meta.xlsx`)

## 4.1 Structure

```{r}
#| label: meta-structure
meta <- read_excel("YvO_meta.xlsx")

cat(sprintf("Samples: %d rows × %d columns\n", nrow(meta), ncol(meta)))
```

| Column | Type | Description |
|---------------------|------------------|----------------------------------|
| `Col_ID` | Character | Unique sample identifier matching raw data columns |
| `Subject_ID` | Character | Subject identifier (**not globally unique** — see note below) |
| `Group` | Factor | Young or Old |
| `Timepoint` | Factor | Pre or Post |
| `Group_Time` | Factor | Young_Pre, Young_Post, Old_Pre, Old_Post |
| `Age` | Numeric | Age in years at enrollment |
| `Sex` | Character | All male ("M") in this cohort |
| `BMI` | Numeric | Body mass index (kg/m²) |
| `DXA_LBM_kg` | Numeric | Dual X-ray absorptiometry lean body mass |
| `VL_thick_cm` | Numeric | Vastus lateralis muscle thickness (cm) |
| `Type_I_fCSA` | Numeric | Type I fiber cross-sectional area (μm²) |
| `Type_II_fCSA` | Numeric | Type II fiber cross-sectional area (μm²) |
| `deadlift_1rm_kg` | Numeric | 1-repetition maximum deadlift (kg) |
| `Total_Training_Volume_kg` | Numeric | Cumulative training volume |

## 4.2 Design Validation

```{r}
#| label: design-validation

# Sample counts per group
cat("=== Samples per Group × Timepoint ===\n")
print(table(meta$Group, meta$Timepoint))

# Subject counts
meta$subject_key <- sub("_(Pre|Post)$", "", meta$Col_ID)
n_young <- length(unique(meta$subject_key[meta$Group == "Young"]))
n_old   <- length(unique(meta$subject_key[meta$Group == "Old"]))

cat(sprintf("\nUnique subjects (by Col_ID prefix): Young = %d, Old = %d\n",
            n_young, n_old))

# Verify pairing
key_counts <- table(meta$subject_key)
cat(sprintf("All subjects have exactly 2 samples (Pre+Post): %s\n",
            all(key_counts == 2)))

# Age ranges
cat(sprintf("\nAge range — Young: %d–%d yr, Old: %d–%d yr\n",
            min(meta$Age[meta$Group == "Young"]),
            max(meta$Age[meta$Group == "Young"]),
            min(meta$Age[meta$Group == "Old"]),
            max(meta$Age[meta$Group == "Old"])))
```

## 4.3 Critical Note: Subject_ID Ambiguity

```{r}
#| label: subject-id-warning

# Find Subject_IDs shared between groups
young_sids <- unique(meta$Subject_ID[meta$Group == "Young"])
old_sids   <- unique(meta$Subject_ID[meta$Group == "Old"])
shared <- intersect(young_sids, old_sids)

cat(sprintf("Subject_IDs shared between Young and Old: %d\n", length(shared)))
cat("Shared IDs:", paste(shared, collapse = ", "), "\n")

cat("\n Example:")
# Show that these are different people
for (s in shared[1:2]) {
  cat(sprintf("\n  %s in Young: age = %d\n", s,
              unique(meta$Age[meta$Subject_ID == s & meta$Group == "Young"])))
  cat(sprintf("  %s in Old:   age = %d\n", s,
              unique(meta$Age[meta$Subject_ID == s & meta$Group == "Old"])))
}

# Also within-group duplicates
within_old <- meta %>% filter(Group == "Old") %>%
  group_by(Subject_ID, Timepoint) %>%
  filter(n() > 1) %>%
  ungroup()
if (nrow(within_old) > 0) {
  cat(sprintf("\nWithin-group (Old) Subject_ID duplicates: %s\n",
              paste(unique(within_old$Subject_ID), collapse = ", ")))
  cat("(These are different individuals from supplement vs placebo arms)\n")
}
```

**Implication for the pipeline:** The repeated-measures blocking factor for `limma::duplicateCorrelation()` must use a **prefix-based subject key** (derived from `Col_ID` by removing the `_(Pre|Post)` suffix) rather than `Subject_ID` alone. Using `Subject_ID` would incorrectly group 4 samples from 2 different individuals under the same block, biasing the within-subject correlation estimate and all downstream test statistics.

The correct subject key is:

``` r
subject_key <- sub("_(Pre|Post)$", "", Col_ID)
# e.g., "OP_S06" and "O_S06" are correctly treated as separate subjects
```

------------------------------------------------------------------------

# 5. HPA Skeletal Muscle Annotations

## 5.1 Purpose

The Human Protein Atlas (HPA) subcellular location and tissue-specific expression annotations are used in the **normalization phase** to filter proteins unlikely to be genuinely present in skeletal muscle (e.g., liver- or brain-specific secreted proteins detected as contaminants).

**Reference:** Uhlén et al. 2015, Science 347:1260419 — tissue-based map of the human proteome

## 5.2 Structure

```{r}
#| label: hpa-structure
hpa <- read.delim("HPA_skeletal_muscle_annotations.tsv")

cat(sprintf("Rows: %d proteins | Columns: %d\n", nrow(hpa), ncol(hpa)))
```

| Column | Description |
|----------------------------|--------------------------------------------|
| `Gene` | HGNC gene symbol (links to raw data `gene` column) |
| `Gene.synonym` | Alternative gene names |
| `Ensembl` | Ensembl gene ID |
| `Uniprot` | UniProt accession (links to raw data `uniprot_id`) |
| `Protein.class` | Functional classification (enzyme, transporter, etc.) |
| `Biological.process` | GO biological process annotations |
| `Molecular.function` | GO molecular function annotations |
| `Subcellular.main.location` | Primary subcellular compartment (used for GO:CC filtering) |

The normalization script uses the `Subcellular.main.location` field to flag and optionally remove proteins annotated to non-muscle-relevant compartments.

------------------------------------------------------------------------

# 6. Data Flow Into the Pipeline

```         
00_input/
├── YvO_raw.xlsx              ──→  01_normalization (raw intensities)
├── YvO_meta.xlsx             ──→  01_normalization (design + demographics + phenotypes)
└── HPA_skeletal_muscle_annotations.tsv  ──→  01_normalization (tissue filter)
```

## 6.1 Downstream Dependencies

| Consumer | Input File | What It Uses |
|---------------------|-----------------------|-----------------------------|
| 01_normalization | `YvO_raw.xlsx` | Protein intensities for filtering + normalization |
| 01_normalization | `YvO_meta.xlsx` | Sample design for group-aware missingness filtering |
| 01_normalization | `HPA_*.tsv` | Skeletal muscle, tissue-relevant, 2filtering |
| 02_Imputation | Normalized DAList (`.rds`) | Output of normalization, not directly from 00_input |
| 03_DEP | Normalized + imputed data | Uses metadata stored in the DAList |
| F6 panels | `YvO_meta.xlsx` (via DAList) | Phenotype scores carried through metadata |

------------------------------------------------------------------------

# 7. Validation Summary

```{r}
#| label: validation-summary

checks <- data.frame(
  Check = c(
    "Raw data dimensions",
    "All meta Col_IDs match raw columns",
    "Balanced design (Pre = Post per group)",
    "Subject_key uniquely pairs Pre+Post",
    "No missing values in design columns",
    "Age groups non-overlapping",
    "HPA annotations available"
  ),
  Status = c(
    sprintf("%d proteins × %d samples", nrow(raw), ncol(raw) - 5),
    sprintf("%d Col_IDs matched", nrow(meta)),
    "Young 17+17, Old 15+15",
    sprintf("%d unique subject_keys, all with 2 samples",
            length(unique(meta$subject_key))),
    "No NAs in Col_ID, Subject_ID, Group, Timepoint, Group_Time",
    sprintf("Young %d–%d yr, Old %d–%d yr",
            min(meta$Age[meta$Group == "Young"]),
            max(meta$Age[meta$Group == "Young"]),
            min(meta$Age[meta$Group == "Old"]),
            max(meta$Age[meta$Group == "Old"])),
    sprintf("%d gene annotations loaded", nrow(hpa))
  )
)
kable(checks, caption = "Input data validation checklist")
```

## 7.1 Known Issues to Address in Downstream Phases

1.  **Subject_ID ambiguity**: 8 Subject_IDs are shared between Young and Old groups (different individuals). The DEP analysis **must** use `subject_key = sub("_(Pre|Post)$", "", Col_ID)` for the `duplicateCorrelation()` blocking factor, **not** `Subject_ID`.

2.  **Pheno file contains summary rows**: `YvO_pheno_calc.xlsx` contains mean, SD, and T-test rows mixed with individual data, and is **not currently consumed** by any downstream stage (01–04). Retained as a reference file for future phenotype work; if adopted in the pipeline, filter summary rows before per-subject analysis.

3.  **Supplement arm**: The Col_ID prefix encodes a supplement condition (O vs OP, Y vs YP). The current pipeline groups these together (Young vs Old), treating supplement status as a nuisance variable. This is acceptable if supplement effects are small relative to age/training effects, but should be documented.

------------------------------------------------------------------------

# 8. Session Information

```{r}
#| label: session-info
sessionInfo()
```
