---
title: "YvO Proteomics Pipeline — Sensitivity Analysis"
author: "DTL"
date: "2026-02-27"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# 1. Overview

This analysis tests whether downstream differential expression results are
robust to the choice of normalization and imputation method. The top 3
normalization methods and top 3 imputation methods (plus no imputation) are
crossed in a 3 x 4 factorial grid, yielding 12 pipeline combinations that
are each run through the same limma model used in `03_DEP`.

**Rationale:** No single normalization or imputation method is universally
optimal for label-free proteomics (Valikangas et al. 2018; Lazar et al. 2016).
A sensitivity analysis quantifies how much the choice matters for the specific
dataset at hand, providing confidence that biological conclusions are not
artifacts of a particular preprocessing decision.

**Input:**

| File | Source |
|------|--------|
| `01_DAList_prenorm.rds` | 01_normalization |
| `03_benchmark_summary.csv` | 02_Imputation |
| `02_mar_mnar_classification.csv` | 02_Imputation |

**Script:** `03_DEP/sensitivity_analysis/a_script/YvO_sensitivity_analysis.R`

---

# 2. Methods

## 2.1 Pipeline Grid

The top 3 normalization methods are selected by composite rank of PCV, PMAD,
and intraclass correlation (PRONE framework; Arend et al. 2025). The top 3
imputation methods are selected by mean NRMSE from the masked-value benchmark
(Jin et al. 2021).

```{r}
#| label: grid-design
norms <- read.csv("../c_data/01_sensitivity_comparison.csv")
cat(sprintf("Normalization methods: %s\n",
            paste(unique(norms$norm), collapse = ", ")))
cat(sprintf("Imputation methods: %s\n",
            paste(unique(norms$imputation), collapse = ", ")))
cat(sprintf("Total pipelines: %d\n", nrow(norms)))
```

A "none" arm (no imputation) is included as a baseline, since the primary
DEP analysis uses non-imputed data with limma handling NAs per-protein
(Karpievitch et al. 2019).

## 2.2 Concordance Metrics

Three complementary metrics assess pipeline agreement:

1. **Pairwise Spearman rho of t-statistics** — rank-order concordance across
   all 12 pipelines for each contrast. Spearman (rather than Pearson) is used
   because it captures monotonic agreement without assuming linearity
   (Valikangas et al. 2018).

2. **Concordance At the Top (CAT) curves** — the fraction of overlap between
   each pipeline's top-k ranked proteins and a consensus ranking, for
   increasing k (Irizarry et al. 2005). The consensus is computed as the mean
   rank (Borda count) across all 12 pipelines.

3. **DEP counts at multiple significance thresholds** — nominal p < 0.05,
   FDR < 0.05, FDR < 0.10, and Pi-score < 0.05 (Xiao et al. 2014). The
   Pi-score (Pi = p^|logFC|) combines statistical significance with effect
   size; values < 0.05 require both.

## 2.3 Contrasts Tested

Four contrasts are tested (matching `03_DEP`):

| Contrast | Definition |
|----------|-----------|
| Training_Young | Young_Post - Young_Pre |
| Training_Old | Old_Post - Old_Pre |
| Aging | Old_Pre - Young_Pre |
| Interaction | (Old_Post - Old_Pre) - (Young_Post - Young_Pre) |

The Reversal contrast is excluded because it is a linear combination of
Aging and Training_Old, making it non-independent.

## 2.4 Limma Model

Each pipeline combination uses the same limma specification as `03_DEP`:
cell-means parameterization (`~ 0 + group`),
`duplicateCorrelation()` for within-subject blocking, and
`eBayes(robust = TRUE)` — mirroring proteoDA's `fit_limma_model()`.

---

# 3. Results

## 3.1 Summary Table

```{r}
#| label: summary-table
summary_df <- read.csv("../c_data/01_sensitivity_comparison.csv")
knitr::kable(summary_df, digits = 4,
             caption = "Pipeline sensitivity comparison (sorted by mean Spearman rho)")
```

## 3.2 Spearman Concordance

```{r}
#| label: spearman-stats
sp <- read.csv("../c_data/02_pairwise_spearman.csv")
overall <- mean(sp$spearman_rho, na.rm = TRUE)
by_contrast <- aggregate(spearman_rho ~ contrast, sp, mean)

cat(sprintf("Overall mean Spearman rho: %.4f\n", overall))
knitr::kable(by_contrast, digits = 4, col.names = c("Contrast", "Mean rho"),
             caption = "Mean pairwise Spearman rho by contrast")
```

## 3.3 DEP Counts

```{r}
#| label: dep-counts
dep <- read.csv("../c_data/03_dep_counts.csv")
dep_wide <- dep |>
  dplyr::filter(criterion == "FDR_05") |>
  dplyr::select(method, contrast, n_total) |>
  tidyr::pivot_wider(names_from = contrast, values_from = n_total)
knitr::kable(dep_wide, caption = "FDR < 0.05 DEP counts per pipeline")
```

---

# 4. Interpretation

```{r}
#| label: interpretation
cat(sprintf("Overall mean Spearman rho: %.4f\n", overall))
if (overall > 0.95) {
  cat("Methods agree closely (rho > 0.95). Pipeline choice has minimal impact.\n")
} else if (overall > 0.85) {
  cat("Methods mostly agree (rho > 0.85). Some variation in extreme ranks.\n")
} else {
  cat("Methods diverge meaningfully (rho < 0.85). Results depend on pipeline.\n")
}
```

**Key findings:**

- The non-imputed pipelines and bpca-imputed pipelines cluster together with
  high concordance (rho > 0.94).
- The `mix_bpca_MinProb` pipelines show systematically lower concordance
  (rho ~ 0.77), driven by the MinProb MNAR imputation introducing left-shifted
  values that alter the ranking of low-abundance proteins.
- The Aging contrast is most robust across pipelines (highest mean rho),
  while the Interaction contrast is most sensitive (lowest mean rho).
- DEP counts are stable across the top pipelines: Aging consistently yields
  400-700+ FDR-significant proteins; Training_Old consistently yields 0-4.

---

# 5. Reports

| File | Location | Contents |
|------|----------|----------|
| `01_sensitivity_comparison.pdf` | b_reports | Spearman heatmaps, CAT curves, DEP count plots |
| `01_sensitivity_comparison.csv` | c_data | Summary table (ranked by mean Spearman rho) |
| `02_pairwise_spearman.csv` | c_data | All pairwise rho values |
| `03_dep_counts.csv` | c_data | DEP counts by method, contrast, and criterion |
| `04_cat_curves.csv` | c_data | CAT curve data (k, concordance, method, contrast) |
| `05_sensitivity_results.xlsx` | c_data | Combined Excel workbook (all data) |
| `06_sensitivity_supplementary.xlsx` | c_data | Manuscript supplementary tables (S1–S6) |

---

# 6. References

- Arend N et al. (2025). PRONE: Normalization evaluation for proteomics.
  *Brief Bioinform* 26(3):bbaf201.
- Irizarry RA et al. (2005). Multiple-laboratory comparison of microarray
  platforms. *Nat Methods* 2(5):345-350.
- Jin L et al. (2021). A comparative study of evaluating missing value
  imputation methods in label-free proteomics. *Sci Rep* 11:1556.
- Karpievitch YV et al. (2019). A statistical framework for protein
  quantitation in bottom-up MS-based proteomics. *BMC Bioinform* 20:391.
- Lazar C et al. (2016). Accounting for the multiple natures of missing
  values in label-free quantitative proteomics. *J Proteome Res* 15(4):1116-1125.
- Valikangas T et al. (2018). A systematic evaluation of normalization
  methods in quantitative label-free proteomics. *Brief Bioinform* 19(1):1-11.
- Webb-Robertson BJ et al. (2015). Review of missing value imputation for
  mass spectrometry-based label-free global proteomics. *J Proteome Res*
  14(3):920-930.
- Xiao Y et al. (2014). A novel significance score for gene selection and
  ranking. *Bioinformatics* 30(6):801-807.

---

# 7. Session Information

```{r}
#| label: session-info
sessionInfo()
```
